% Exercise boxes
\mdtheorem[style=theoremstyle,frametitlebackgroundcolor=black,frametitlefontcolor=white]{exercise}{Exercise}

% Try to find better workaround for \sloppy (cf https://tex.stackexchange.com/q/153872)

\newcommand{\llpr}[1]{{\llparenthesis #1 \rrparenthesis}}
\newcommand{\FamSet}{{\mathsf{Fam}(\mathsf{Set})}}
\newcommand{\Alg}{\mathsf{Alg}}
\newcommand{\inl}{\mathsf{inl}}
\newcommand{\inr}{\mathsf{inr}}
\newcommand{\soltitle}[1]{\vspace{12pt}\noindent\textbf{Solution \ref{#1}:}}

\chapter{Type Refinements as Indexed Inductive Types}

In the course of writing code in a simply typed language, one may find oneself wishing to constrain
the values of a type based on some ``feature'' of its values. For instance, when writing functions
over lists or trees, it may be desirable to enforce constraints on the length or height of the
input, or to ensure that the values it stores conform to some precondition. A classic example is
the list \texttt{zip} operation: the author of this function may wish to enforce that its arguments
have the same length.

\emph{Refinement types} enrich simple type systems to allow programmers to express such constraints.
For example, the type of lists of length \texttt{n} can be written
\[ \texttt{\{xs : List A | length(xs) = n\}} \]

In dependently typed languages, such refinement types can be na\"ively translated as dependent
pairs, attaching to lists proofs that they satisfy the length constraint:

\[ \sum_{xs ~:~ \texttt{List}~\alpha} \texttt{length}(xs) = n \]

However, dependent types permit a more elegant encoding: such constraints can be directly ``baked
into'' the definition of an inductive type via type indices, ensuring that all values inhabiting the
type are correct by construction. Our example above, for instance, yields the type of length-indexed
vectors:\footnote{Throughout this chapter, we will write code snippets in a syntax similar to that
of Lean 4, though we assume all constructors live in the global namespace. Those familiar with Agda
can read \texttt{inductive} as \texttt{data} and \texttt{Type} as \texttt{Set}.}

\begin{lstlisting}
inductive Vector (A : Type) : Nat -> Type
  | vnil  : Vector A 0
  | vcons : {n : Nat} -> A -> Vector A n -> Vector A (succ n)
\end{lstlisting}

The type \texttt{Vector} naturally internalizes the \texttt{length} refinement from the simply typed
setting. In this chapter, we will use a categorical construction to generalize this transformation
to arbitrary inductive types and a broad class of refinements. In other words, we answer the
following question: how can a refinement of a type in a simply typed language be translated into an
indexed inductive type in a dependently typed one?

Although this question serves as our motivation, the actual aim of this chapter is threefold:
\begin{itemize}
  \item To give an account of a categorical semantics of inductive and indexed inductive types;
  \item To illustrate an application of these tools in the form of the translation discussed above;
  \item To provide motivating intuition for the study of \emph{fibrations}, which we regrettably do
        not have space to treat herein.
\end{itemize}

The content of this chapter is largely based on the work of Atkey, Johann, and
Ghani~\citep{atkey2011refining}. The content of the first section on inductive types also draws on
Wadler's ``Recursive Types for Free!''~\citep{wadler1990recursive} and scribed lecture notes from
New's presentation of similar material \citep{new2023category}.

\section{Categorifying Inductive Types}

We will begin by adding a new programming-language feature to our categorical repertoire:
definitions of inductive datatypes.

\subsection{Wherefore More Constructions?}

Let us first consider why the facilities afforded to us by our existing semantic development are
inadequate to express inductive type definitions in full generality. This may not be immediately
obvious: for instance, as we have previously seen, the datatype of booleans
\begin{lstlisting}
inductive Bool : Type where
  | true  : Bool
  | false : Bool
\end{lstlisting}
has elements isomorphic to those of $\texttt{Unit} \oplus \texttt{Unit}$, which we interpret
categorically as $1 \oplus 1$. We can even model finite parameterized datatypes: the option datatype
\begin{lstlisting}
inductive Option (A : Type) : Type where
  | some : A -> Option A
  | none : Option A
\end{lstlisting}
has elements isomorphic to those of $\texttt{A} \oplus \texttt{Unit}$, which we interpret as
$\llbr{\texttt{A}} \oplus 1$ (assuming that we are able to interpret \texttt{A}). More generally, to
interpret the elements of a given inductive type, we can treat each constructor argument as a term
in a product, then sum together the representations of each constructor. Where lies the problem?

You may notice that our wording above---interpreting ``elements of the type'' induced by the
definition, rather than the ``definition'' itself---is somewhat roundabout: this is intentional. An
inductive type is characterized not merely by its ground inhabitants, but also by the \emph{maps
into} the type (i.e., its constructor forms). An inductive type also comes equipped with a
\emph{recursor} (also known as an \emph{iterator} or \emph{fold}) that admits recursive definitions
over the type. The account above, which describes only the object denoted by the type itself, does
not directly characterize these facets of an inductive definition.

Moreover---and more problematically---we also currently lack the facilities to provide a general
categorical account of infinite inductive types, such as the type of natural numbers
\begin{lstlisting}
inductive Nat : Type where
  | zero : Nat
  | succ : Nat -> Nat\end{lstlisting}
or the type of lists
\begin{lstlisting}
inductive List (A : Type) : Type where
  | nil  : List A
  | cons : A -> List A -> List A
\end{lstlisting}
since the tools we have previously used to give the semantics of lambda calculi---(finite)
(co)products, exponential objects, and initial and terminal objects---do not (at least
straightforwardly)\footnote{We would prefer a more ``natural'' account of inductive types than, say,
Church encodings.} admit the construction of sums and products of unbounded arity. To give a
complete account of general inductive definitions, we will need to develop some new (and repurpose
some old) tools.

We will motivate our translation of inductive datatypes using our standard approach to translating
PL concepts into category theory: translating inference rules. Specifically, we will translate the
inference rules for booleans and natural numbers into categorical constructions, then observe some
patterns that will allow us to generalize these translations to arbitrary inductive types.

\subsection{Categorifying Booleans}

To begin, we return to the example of the type \texttt{Bool}. Its constructors are characterized by
the following typing rules:

\begin{mathpar}
\inferrule*[right=T-true]{~}{\vdash \texttt{true} : \texttt{Bool}}

\inferrule*[right=T-false]{~}{\vdash \texttt{false} : \texttt{Bool}}
\end{mathpar}

Note that we write these rules with an empty context, rather than an arbitrary $\Gamma$. Recalling
that the empty context is represented by the terminal object $1$, these rules correspond to the
existence of two \emph{global elements} $\texttt{true}, \texttt{false} : 1 \to \llbr{\texttt{Bool}}$.

\begin{definition}[Global element]
  In a category with terminal objects, a \emph{global element} of an object $A$ is a morphism
  $1 \to A$. It is the categorical interpretation of an ``element'' of an object.
\end{definition}

Here, our use of global elements corresponds to the fact that \texttt{true} and \texttt{false} are
distinguished members of the ``collection'' denoted by \texttt{Bool}. The rules above translate into
the following diagram:

% https://q.uiver.app/#q=WzAsMyxbMCwwLCIxIl0sWzIsMCwiXFx0ZXh0dHR7Qm9vbH0iXSxbNCwwLCIxIl0sWzAsMSwiXFx0ZXh0dHR7dHJ1ZX0iXSxbMiwxLCJcXHRleHR0dHtmYWxzZX0iLDJdXQ==
\[\begin{tikzcd}
	1 && {\llbr{\texttt{Bool}}} && 1
	\arrow["{\texttt{true}}", from=1-1, to=1-3]
	\arrow["{\texttt{false}}"', from=1-5, to=1-3]
\end{tikzcd}\]

However, we want to be able to map \emph{out} of our inductive types as well. To do so, we must
define the recursor for the Boolean type, canonically known as \texttt{if}:\footnote{We will adopt
the convention of writing recursors with their \emph{major premise} (i.e., the value being
``recursed on'') as the final argument. This is opposite the usual notation in functional
programming but will better agree with the categorical formulation.}

\begin{mathpar}
\inferrule*[right=T-if]{\vdash t : \tau \\ \vdash e : \tau}{b : \texttt{Bool} \vdash \texttt{if}(t, e)(b) : \tau}
\end{mathpar}

The recursor is operationally characterized by $\beta$ and $\eta$ laws. The $\beta$ laws say that
\texttt{if} should ``case correctly'': it should evaluate to its ``then'' branch on a \texttt{true}
input, and to its ``else'' branch on a \texttt{false} one. The $\eta$ law, meanwhile, says that
``things that behave like \texttt{if}s are judgmentally equal to \texttt{if}s;'' one can think of
this as a restricted form of function extensionality:

\begin{mathpar}
\inferrule*[right=$\beta^{\texttt{if}}_1$]
  {\vdash t : \tau \\ \vdash e : \tau}{\vdash \texttt{if}(t, e)(\texttt{true}) \equiv t : \tau}

\inferrule*[right=$\beta^{\texttt{if}}_2$]
  {\vdash t : \tau \\ \vdash e : \tau}{\vdash \texttt{if}(t, e)(\texttt{false}) \equiv e : \tau}

\inferrule*[right=$\eta^{\texttt{if}}$]
  {x : \texttt{Bool} \vdash d : \tau \quad
   \vdash d[\texttt{true}/x] \equiv t : \tau \quad
   \vdash d[\texttt{false}/x] \equiv e : \tau}
  {x : \texttt{Bool} \vdash d \equiv \texttt{if}(t,e)(x) : \tau}
\end{mathpar}

As in our previous translations of equational laws, the $\beta$ laws correspond to equalities of
morphisms, while the $\eta$ law is a uniqueness criterion. In particular, the $\beta$ laws require
that if we have morphisms $t, e : 1 \to \llbr{\tau}$, then $\texttt{if}(t,e) \circ \texttt{true} =
t$ and $\texttt{if}(t,e) \circ \texttt{false} = e$. Meanwhile, the $\eta$ law says that if we have
some expression $d$ such that it can be substituted for $\texttt{if}(t, e)$ in the preceding two
equalities\footnote{Recall that we interpret substitution as composition of morphisms.}---i.e., $d
\circ \texttt{true} \equiv t$ and $d \circ \texttt{false} \equiv e$---then it is equal to the
recursor: in other words, the recursor denotes the unique morphism satisfying these equalities.
Putting these together, we obtain the following pair of commutative triangles:

% https://q.uiver.app/#q=WzAsNCxbMCwwLCIxIl0sWzIsMCwiXFx0ZXh0dHR7Qm9vbH0iXSxbNCwwLCIxIl0sWzIsMiwiXFx0YXUiXSxbMCwxLCJcXHRleHR0dHt0cnVlfSJdLFsyLDEsIlxcdGV4dHR0e2ZhbHNlfSIsMl0sWzEsMywiXFx0ZXh0dHR7aWZ9KHQsZSkiLDJdLFswLDMsInQiLDFdLFsyLDMsImUiLDFdXQ==
\[\begin{tikzcd}
	1 && {\llbr{\texttt{Bool}}} && 1 \\
	\\
	&& \llbr{\tau}
	\arrow["{\texttt{true}}", from=1-1, to=1-3]
	\arrow["t"{description}, from=1-1, to=3-3]
	\arrow["{\texttt{if}(t,e)}"', from=1-3, to=3-3, dashed]
	\arrow["{\texttt{false}}"', from=1-5, to=1-3]
	\arrow["e"{description}, from=1-5, to=3-3]
\end{tikzcd}\]

For reasons that will become clear shortly, it is helpful to expand these triangles into squares:

% https://q.uiver.app/#q=WzAsNixbMCwwLCIxIl0sWzIsMCwiXFx0ZXh0dHR7Qm9vbH0iXSxbNCwwLCIxIl0sWzIsMiwiXFx0YXUiXSxbMCwyLCIxIl0sWzQsMiwiMSJdLFswLDEsIlxcdGV4dHR0e3RydWV9Il0sWzIsMSwiXFx0ZXh0dHR7ZmFsc2V9IiwyXSxbMSwzLCJcXHRleHR0dHtpZn0odCxlKSIsMl0sWzQsMywidCIsMV0sWzUsMywiZSIsMV0sWzIsNSwiXFxtYXRoc2Z7aWR9XzEiLDFdLFswLDQsIlxcbWF0aHNme2lkfV8xIiwxXV0=
\begin{equation}
\begin{tikzcd}
	1 && {\llbr{\texttt{Bool}}} && 1 \\
	\\
	1 && \llbr{\tau} && 1
	\arrow["{\texttt{true}}", from=1-1, to=1-3]
	\arrow["{\mathsf{id}_1}"{description}, from=1-1, to=3-1]
	\arrow["{\texttt{if}(t,e)}"', from=1-3, to=3-3, dashed]
	\arrow["{\texttt{false}}"', from=1-5, to=1-3]
	\arrow["{\mathsf{id}_1}"{description}, from=1-5, to=3-5]
	\arrow["t"{description}, from=3-1, to=3-3]
	\arrow["e"{description}, from=3-5, to=3-3]
\end{tikzcd}
\label{fig:bool-squares}
\end{equation}

\begin{exercise}
\label{ex:denote-bool}
\sloppy
Verify that $1 \oplus 1$ is a suitable choice for $\llbr{\texttt{Bool}}$. What are the denotations
of \texttt{true}, \texttt{false}, and $\texttt{if}(t,e)$?
\end{exercise}

\subsection{Categorifying Natural Numbers}

We now perform a similar translation for natural numbers. The type \texttt{Nat} has two
constructors, \texttt{zero} and \texttt{succ}, and is equipped with a recursor $\texttt{rec}(e_z,
x.e_s)(n)$. Notice that, unlike \texttt{if}, the recursor for \texttt{Nat} binds a variable in the
successor case: this will be the recursive result of evaluating the recursor on the substructure of
$n$. The typing rules are given below:

\begin{mathpar}
\inferrule*[right=T-z]{~}{\vdash \texttt{zero} : \texttt{Nat}}

\inferrule*[right=T-s]{~}{n : \texttt{Nat} \vdash \texttt{succ}(n) : \texttt{Nat}}

\inferrule*[right=T-natrec]
  {\vdash e_z : \tau \\ x : \texttt{Nat} \vdash e_s : \tau}
  {n : \texttt{Nat} \vdash \texttt{rec}_\texttt{Nat}(e_z, x. e_s)(n) : \tau}
\end{mathpar}

The $\beta$ rules for the recursor express its behavior as a fold over the structure of a natural
number: applying the recursor to zero evaluates to the first arm, while applying it to a successor
binds $x$ to the recursive result of folding over its predecessor, then evaluates $e_s$:
\begin{mathpar}
\inferrule*[right=$\beta^{\texttt{rec}_\texttt{Nat}}_z$]
  {\vdash e_z : \tau \\ x : \texttt{Nat} \vdash e_s : \tau}
  {\vdash \texttt{rec}_{\texttt{Nat}}(e_z, x. e_s)(\texttt{zero}) \equiv e_z : \tau}

\inferrule*[right=$\beta^{\texttt{rec}_\texttt{Nat}}_s$]
  {\vdash e_z : \tau \\ x : \texttt{Nat} \vdash e_s : \tau \\ \vdash n : \texttt{Nat}}
  {\vdash \texttt{rec}_\texttt{Nat}(e_z, x. e_s)(\texttt{succ}(n)) \equiv e_s[\texttt{rec}_\texttt{Nat}(e_z, x. e_s)(n)/x] : \tau}
\end{mathpar}

Finally, the $\eta$ rule says that any expression $e$ that satisfies the above equations is equal to
the corresponding recursor application:

\begin{mathpar}
\inferrule*[right=$\eta^{\texttt{rec}_\texttt{Nat}}$]
  {y : \texttt{Nat} \vdash e : \tau \\ \vdash e[\texttt{zero}/y] \equiv e_z : \tau \\ n : \texttt{Nat} \vdash e[\texttt{succ}(n)/y] \equiv e_s[e/x] : \tau }
  {y : \texttt{Nat} \vdash e \equiv \texttt{rec}_\texttt{Nat}(e_z, x.e_s)(y) : \tau}
\end{mathpar}

As with Booleans, categorifying the rules for natural numbers yields an object representing the type
\texttt{Nat}, a morphism for each constructor, and, for each pair of ``combining-function''
morphisms running ``parallel to'' the constructors, a unique morphism denoted by the
recursor.\footnote{Note that we obtain from the $\eta$ law the uniqueness condition---that any
morphism $e$ satisfying the $\beta$-derived equalities $e \circ \texttt{zero} = e_z$ and $e \circ
\texttt{succ} = e_s$ is equal to the recursor morphism---by recalling that substitution is
interpreted as composition.} In the case of natural numbers, however, the resulting construction has
a special name: the \emph{natural numbers object}.

\begin{definition}[Natural numbers object]
\sloppy
  A \emph{natural numbers object} in a Cartesian closed category $\calC$ consists of
  \begin{itemize}
    \item An object $N$ of $\calC$ and
    \item Morphisms $z : 1 \to N$ and $s : N \to N$
  \end{itemize}
  such that, for every object $A$ of $\calC$ and pair of morphisms $f_z : 1 \to A$ and $f_s : A \to
  A$, there exists a unique morphism $u$ such that the following diagram commutes:
  % https://q.uiver.app/#q=WzAsNSxbMCwwLCIxIl0sWzIsMCwiTiJdLFs0LDAsIk4iXSxbMiwyLCJBIl0sWzQsMiwiQSJdLFswLDEsInoiXSxbMSwzLCJ1IiwyXSxbNCwzLCJlX3MiLDFdLFsyLDQsInUiLDFdLFsyLDEsInMiLDJdLFswLDMsImVfeiIsMV1d
  \[\begin{tikzcd}
    1 && N && N \\
    \\
    && A && A
    \arrow["z", from=1-1, to=1-3]
    \arrow["{e_z}"{description}, from=1-1, to=3-3]
    \arrow["u"', from=1-3, to=3-3, dashed]
    \arrow["s"', from=1-5, to=1-3]
    \arrow["u"{description}, from=1-5, to=3-5, dashed]
    \arrow["{e_s}"{description}, from=3-5, to=3-3]
  \end{tikzcd}\]
\end{definition}

Here, $u$ is $\texttt{rec}_\texttt{Nat}(e_z, x.e_s)$. As before, it will be helpful to instead view this
diagram as a pair of squares:

% https://q.uiver.app/#q=WzAsNyxbMCwwLCIxIl0sWzIsMCwiXFxsbGJye1xcdGV4dHR0e05hdH19Il0sWzQsMCwiXFxsbGJye1xcdGV4dHR0e05hdH19Il0sWzIsMiwiXFxsbGJye1xcdGF1fSJdLFswLDIsIjEiXSxbNCwyLCJcXGxsYnJ7XFx0YXV9Il0sWzMsMF0sWzAsMSwiXFx0ZXh0dHR7en0iXSxbMSwzLCJcXHRleHR0dHtyZWN9KGVfeix4Ln5lX3MpIiwyXSxbNCwzLCJlX3oiLDFdLFs1LDMsImVfcyIsMV0sWzIsNSwiXFx0ZXh0dHR7cmVjfShlX3oseC5+ZV9zKSIsMV0sWzAsNCwiXFxtYXRoc2Z7aWR9XzEiLDFdLFsyLDEsIlxcdGV4dHR0e3N9IiwyXV0=
\begin{equation}
\begin{tikzcd}
	1 && {\llbr{\texttt{Nat}}} & {} & {\llbr{\texttt{Nat}}} \\
	\\
	1 && {\llbr{\tau}} && {\llbr{\tau}}
	\arrow["{\texttt{zero}}", from=1-1, to=1-3]
	\arrow["{\mathsf{id}_1}"{description}, from=1-1, to=3-1]
	\arrow["{\texttt{rec}_\texttt{Nat}(e_z,x.e_s)}"', from=1-3, to=3-3, dashed]
	\arrow["{\texttt{succ}}"', from=1-5, to=1-3]
	\arrow["{\texttt{rec}_\texttt{Nat}(e_z,x.e_s)}"{description}, from=1-5, to=3-5, dashed]
	\arrow["{e_z}"{description}, from=3-1, to=3-3]
	\arrow["{e_s}"{description}, from=3-5, to=3-3]
\end{tikzcd}
\label{fig:nat-squares}
\end{equation}

\begin{exercise}
\label{ex:interp-exnat}
\sloppy
  Give a categorical interpretation of the type of \emph{extended natural numbers}, i.e., the type
  of natural numbers extended with an extra element representing infinity, defined as follows:
\begin{lstlisting}
inductive ExtNat : Type where
  | zero : ExtNat
  | succ : ExtNat -> ExtNat
  | inf  : ExtNat
\end{lstlisting}

  \emph{Hint:} The resulting diagram will look a bit different from the two preceding examples.
\label{ex:ext-nat-interp}
\end{exercise}

\subsection{A General Categorification of Inductive Types}

Let us now take stock of the preceding examples and generalize what we have seen. In each of our
translations above, we did the following:

\begin{itemize}
  \item First, interpret the constructors of an inductive type as morphisms from the interpretation
  of the constructor's domain to the type's denotation. That is, a constructor $\texttt{c} :
  \texttt{D} \to \texttt{T}$ of \texttt{T} becomes a morphism $\llbr{\texttt{D}}
  \xrightarrow{\texttt{c}} \llbr{\texttt{T}}$.
  \item Then, interpret the recursor as a unique morphism $\llbr{\texttt{T}} \to A$ for every
  collection of morphisms into $A$ ``parallel to the constructors,'' i.e., morphisms $t_i$ that map
  into $A$ from the domain of the $i$th constructor where $\llbr{\texttt{T}}$ has been replaced by
  $A$. For instance, if $\texttt{T}$ has two constructors
  \[ \texttt{c}_1 : \texttt{Bool} \to \texttt{T} \quad\text{and}\quad
     \texttt{c}_2 : \texttt{T} \times \texttt{Nat} \to \texttt{T}, \]
  its recursor is interpreted as a unique morphism $u_{f_1, f_2} : \llbr{\texttt{T}} \to A$ for
  all objects $A$ and morphisms
  \[ f_1 : \llbr{\texttt{Bool}} \to A \quad\text{and}\quad
     f_2 : A \times \llbr{\texttt{Nat}} \to A. \]
  Moreover, the resulting squares must commute.
\end{itemize}

This procedure is a useful heuristic but is not yet a precise categorical account of an inductive
type. For instance, the ``replacement'' performed in the second step is merely a syntactic
operation. We will now make this categorical translation more precise.

We start by consolidating our commutative squares. Recall that any inductive type with multiple
constructors can be equivalently expressed as a having a single constructor whose domain is the sum
of the domains of the original constructors. That is, the constructors $c_1 : A \to T$, $c_2 : B \to
T$, and $c_3 : C \to T$ can be equivalently represented by a single constructor $c : A \oplus B
\oplus C \to T$. The recursor is then parameterized by a single morphism running  ``parallel'' to
the constructor that maps into $A$ from an analogous coproduct over $A$.

Applying this rewriting to the \texttt{Nat} datatype depicted in diagram
(\ref{fig:nat-squares}), we obtain
% https://q.uiver.app/#q=WzAsNCxbMCwwLCIxIFxcb3BsdXMgXFxsbGJye1xcdGV4dHR0e05hdH19Il0sWzIsMCwiXFxsbGJye1xcdGV4dHR0e05hdH19Il0sWzAsMiwiMSBcXG9wbHVzIEEiXSxbMiwyLCJBIl0sWzAsMSwiXFxsYW5nbGUgeiwgc1xccmFuZ2xlIl0sWzEsMywiXFxtYXRodHR7aXRlcn0oXFxsYW5nbGUgZl96LCBmX3NcXHJhbmdsZSkiXSxbMiwzLCJcXGxhbmdsZSBmX3osIGZfc1xccmFuZ2xlIiwyXSxbMCwyLCJcXGxlZnRcXGxhbmdsZSBcXG1hdGhzZntpbmx9IFxcY2lyYyBcXG1hdGhzZntpZH1fMSwgXFxtYXRoc2Z7aW5yfSBcXGNpcmMgXFxtYXRoc2Z7aXRlcn0oXFxsYW5nbGUgZl96LCBmX3NcXHJhbmdsZSkgXFxyaWdodFxccmFuZ2xlIiwyXV0=
\[\begin{tikzcd}
	{1 \oplus \llbr{\texttt{Nat}}} && {\llbr{\texttt{Nat}}} \\
	\\
	{1 \oplus A} && A
	\arrow["{[ z, s]}", from=1-1, to=1-3]
	\arrow["{\left[ \inl \mathop\circ \mathsf{id}_1,~ \inr \mathop\circ \texttt{rec}_\texttt{Nat}([ f_z, f_s]) \right]}"', from=1-1, to=3-1]
	\arrow["{\texttt{rec}_\texttt{Nat}([ f_z, f_s])}", from=1-3, to=3-3, dashed]
	\arrow["{[ f_z, f_s]}"', from=3-1, to=3-3]
\end{tikzcd}\]
and a similar transformation occurs for diagram (\ref{fig:bool-squares}). (Above and hereafter, we
use the notation $[f, g]$ to denote the unique morphism obtained from the universal property of
coproducts. Thinking of morphisms as functions, $[f, g]$ does a case split on its input, mapping
elements of the left-hand summand using $f$ and those of the right-hand summand using
$g$.)

The top-left corner should look familiar: it is the body of the recursive type equation that defines
the natural numbers. That is, using a least-fixed-point operator $\mu$, we would write
\[ \texttt{Nat} \triangleq \mu t.~1 \oplus t, \]
so that \texttt{Nat} is the (smallest) type satisfying the type isomorphism
\[ \texttt{Nat} \cong 1 \oplus \texttt{Nat}. \]
Similarly, diagram (\ref{fig:bool-squares}) is ``consolidated'' into a single square with a
constructor with domain $1 \oplus 1$:
\[\begin{tikzcd}
	{1 \oplus 1} && {\llbr{\texttt{Bool}}} \\
	\\
	{1 \oplus 1} && A
	\arrow["{[ z, s]}", from=1-1, to=1-3]
	\arrow["{\left[ \inl \mathop\circ \mathsf{id}_1,~ \inr \mathop\circ \mathsf{id}_1 \right]}"', from=1-1, to=3-1]
	\arrow["{\texttt{if}([ t, e ])}", from=1-3, to=3-3, dashed]
	\arrow["{[ t, e ]}"', from=3-1, to=3-3]
\end{tikzcd}\]
% \begin{exercise}
% \sloppy
%   ``Consolidate'' diagram (\ref{fig:bool-squares}) as we did for diagram (\ref{fig:nat-squares})
%   above, and verify that the (degenerate) recursive type equation for Booleans emerges.
% \end{exercise}

With this consolidation, every inductive type now translates to a single object (the top-left
corner of the square), a single morphism (the constructor), and a universal property that translates
the recursor. The ``shape'' of the objects on the left-hand side of the commutative square are
determined by the ``shape'' of the inductive type. More precisely, using the observation in the
preceding paragraph, that ``shape'' is determined precisely by the expression $\tau$ in the
fixed-point equation $\mu t.~\tau$ for the inductive type.\footnote{You may wonder why we do not
give a general categorification of inductive types simply by translating the general typing rules
for recursive types, as we did for the examples above. Indeed, this is one viable path, but the
relevant abstracted definitions acquire some bureaucratic machinery that we prefer not to deal
with.}

Adopting a categorical lens, we might observe that an expression $t.~\tau$ defines a mapping from a
type $t$ to some type $\tau$ parameterized thereby. In our categorical interpretation, types form a
category; thus, such a mapping from types to types ought to correspond to a functor. More
specifically, since we are mapping from a category (of types) into itself, this is an
\emph{endofunctor}. The endofunctor associated with the type of natural numbers is defined by
\begin{align*}
  F_\texttt{Nat}(X) &= 1 \oplus X\\
  F_\texttt{Nat}(f : X \to Y) &= [\inl \circ \mathsf{id}_1, \inr \circ f] : 1 \oplus X \to 1 \oplus Y
\end{align*}

Of course, the mapping $t.~\tau$ is not itself a type; rather, we apply the least-fixed-point
operator $\mu$ to obtain an inductive type. We should expect a corresponding maneuver in the world
of categories: we interpret an inductive type as the \emph{least fixed point} $\mu F$ of an
endofunctor $F$ on our semantic domain: i.e., the ``least'' object $\mu F$ satisfying $F(\mu F) =
\mu F$.

This naturally invites the following question: in the absence of an \emph{a priori} ordering on
objects, how do we make precise the fact that $\mu F$ should be ``least'' among the fixed points of
$F$? Appealing to our categorical intuition, we should expect this to be some kind of initiality or
colimiting property (recall that, using our order-theoretic lens, we think of initial objects as
``minimal''). And, in fact, as we will now demonstrate, we have already seen such a property in our
Boolean and natural-number examples!

The key is to recall that an inductive type is defined not only by an object but also by a morphism
representing its constructor that tells us how to ``map into'' the type. Thus, instead of merely
considering the ``smallest'' object among the fixed points of $F$, we should instead consider the
``smallest'' \emph{pair} of an $F$-fixed point \emph{and} its corresponding constructor morphism.
Such a pair is an example of an \emph{algebra} over the endofunctor $F$, also known as an
\emph{$F$-algbera}.

\begin{definition}[$F$-Algebra]
\sloppy
  Let $F : \calC \to \calC$ be an endofunctor. An \emph{$F$-algebra}, or \emph{algebra over $F$}, is
  a pair $(A, \alpha)$ where $A$ is an object in $\calC$ and $\alpha : F A \to A$ is a morphism.
  
  $A$ is called the \emph{carrier} of the $F$-algebra, and $\alpha$ is the \emph{structure map}.
  Note that the structure map $\alpha$ alone is sometimes referred to as an $F$-algebra, leaving $A$
  implicit.
\end{definition}

We are looking for the ``smallest'' $F$-algebra $(A, \alpha)$ such that its carrier $A$ is a fixed
point. Let us first, however, consider simply the ``smallest'' $F$-algebra, without any further
restrictions. We do this because it turns out that there is a natural interpretation of ``smallest''
in terms of initiality. In particular, $F$-algebras form a category $\Alg_F$, so the
``smallest'' $F$-algebra is simply the initial object in this category.

\begin{definition}[Category of $F$-algebras]
\label{def:cat-of-algebras}
  Fix an endofunctor $F : \calC \to \calC$. The \emph{category of $F$-algebras} (or \emph{category
  of algebras over $F$}) is defined by the following:
  \begin{itemize}
    \item Objects are $F$-algebras $(A, \alpha : F A \to A)$.
    \item Morphisms from $(A, \alpha)$ to $(B, \beta)$ are morphisms $f : A \to B$ in $\calC$ such
          that the following square commutes:
    % https://q.uiver.app/#q=WzAsNCxbMCwwLCJGIEEiXSxbMiwwLCJBIl0sWzAsMiwiRiBCIl0sWzIsMiwiQiJdLFswLDEsIlxcYWxwaGEiXSxbMSwzLCJmIl0sWzIsMywiXFxiZXRhIiwyXSxbMCwyLCJGIGYiLDJdXQ==
    \[\begin{tikzcd}
      {F A} && A \\
      \\
      {F B} && B
      \arrow["\alpha", from=1-1, to=1-3]
      \arrow["{F f}"', from=1-1, to=3-1]
      \arrow["f", from=1-3, to=3-3]
      \arrow["\beta"', from=3-1, to=3-3]
    \end{tikzcd}\]
    \item Identity and composition are lifted from $\calC$.
  \end{itemize}
\end{definition}

\begin{definition}[Initial $F$-algebra]
\sloppy
  Fix an endofunctor $F : \calC \to \calC$. The \emph{initial $F$-algebra} (or \emph{initial algebra
  over $F$}), if it exists, is the initial object in the category of $F$-algebras.

  We write $\mathsf{in}_F$ for the (structure map of) the initial $F$-algebra. Given another
  $F$-algebra $(A, \alpha)$, we write $\llpr{\alpha}_F$ (or just $\llpr{\alpha}$ if $F$ is clear
  from context) for the unique morphism from the initial object to $(A, \alpha)$.
\end{definition}

We have thus found a suitable notion of a ``smallest'' $F$-algebra. But what about the smallest
$F$-algebra whose carrier is a fixed point of $F$? In fact, these are one and the same (up to
isomorphism): it will always be the case that the carrier of the initial $F$-algebra is (isomorphic
to) a fixed point of $F$. In other words, if $(A, \alpha)$ is the initial $F$-algebra, then $F A
\cong A$. This is the content of \emph{Lambek's lemma}.

\begin{proposition}[Lambek's lemma]
  Let $F : \calC \to \calC$ be an endofunctor. If $(A, \mathsf{in}_F : F A \to A)$ is the initial
  $F$-algebra, then $\mathsf{in}_F$ is an isomorphism.
\end{proposition}

\begin{proof}
  Let $(A, \mathsf{in}_F : F A \to A)$ be the initial $F$-algebra. We will show that $\mathsf{in}_F$
  is an isomorphism by showing that its inverse is $\llpr{F~\mathsf{in}_F} : A \to F A$, the unique
  $F$-algebra morphism $(A, \mathsf{in}_F) \to (F A, F~\mathsf{in}_F)$ induced by initiality.

  First, observe that $\mathsf{in}_F$ is an $F$-algebra morphism $(F A, F~\mathsf{in}_F) \to (A,
  \mathsf{in}_F)$, since the following diagram trivially commutes:

  % https://q.uiver.app/#q=WzAsNCxbMCwwLCJGIChGIEEpIl0sWzIsMCwiRiBBIl0sWzAsMiwiRiBBIl0sWzIsMiwiQSJdLFswLDEsIkYgXFxhbHBoYSJdLFswLDIsIkYgXFxhbHBoYSIsMl0sWzEsMywiXFxhbHBoYSJdLFsyLDMsIlxcYWxwaGEiLDJdXQ==
  \[\begin{tikzcd}
    {F (F A)} && {F A} \\
    \\
    {F A} && A
    \arrow["{F~\mathsf{in}_F}", from=1-1, to=1-3]
    \arrow["{F~\mathsf{in}_F}"', from=1-1, to=3-1]
    \arrow["\mathsf{in}_F", from=1-3, to=3-3]
    \arrow["\mathsf{in}_F"', from=3-1, to=3-3]
  \end{tikzcd}\]

  Thus, we have the following diagram in $\Alg_F$:

  % https://q.uiver.app/#q=WzAsMyxbMCwwLCIoQSwgXFxhbHBoYSkiXSxbMCwyLCIoRiBBLCBGIFxcYWxwaGEpIl0sWzAsNCwiKEEsIFxcYWxwaGEpIl0sWzAsMSwiXFxsbHBye0YgXFxhbHBoYX0iXSxbMSwyLCJcXGFscGhhIl0sWzAsMiwiXFxtYXRoc2Z7aWR9X3soQSwgXFxhbHBoYSl9IiwyLHsiY3VydmUiOjR9XV0=
  \[\begin{tikzcd}
    {(A, \mathsf{in}_F)} \\
    \\
    {(F A, F~\mathsf{in}_F)} \\
    \\
    {(A, \mathsf{in}_F)}
    \arrow["{\llpr{F~\mathsf{in}_F}}", from=1-1, to=3-1]
    \arrow["{\mathsf{id}_A}"', curve={height=36pt}, from=1-1, to=5-1]
    \arrow["\mathsf{in}_F", from=3-1, to=5-1]
  \end{tikzcd}\]
  
  Since $(A, \mathsf{in}_F)$ is initial, there is only one $F$-algebra morphism from $(A,
  \mathsf{in}_F)$ to itself, so this diagram must commute, i.e.,
  \[ \mathsf{in}_F \circ \llpr{F~\mathsf{in}_F} = \mathsf{id}_A.\]
  Accordingly, $\llpr{F~\mathsf{in}_F}$ is a right inverse for $\mathsf{in}_F$.
  
  To show that it is also a left inverse, we apply $F$ to both sides of the preceding equation and
  rewrite using the functor laws:
  \[ F~\mathsf{in}_F \circ F \llpr{F~\mathsf{in}_F} = \mathsf{id}_{F A}. \]
  Then we observe that
  \[ F~\mathsf{in}_F \circ F \llpr{F~\mathsf{in}_F} = \mathsf{in}_F \circ \llpr{F~\mathsf{in}_F} \]
  from the fact that $\llpr{F~\mathsf{in}_F}$ is an $F$-algebra morphism:
  % https://q.uiver.app/#q=WzAsNCxbMCwwLCJGIEEiXSxbMiwwLCJBIl0sWzAsMiwiRiAoRiBBKSJdLFsyLDIsIkYgQSJdLFswLDEsIlxcbWF0aHNme2lufV9GIl0sWzIsMywiRn5cXG1hdGhzZntpbn1fRiJdLFswLDIsIkYgXFxsbHBye0Z+XFxtYXRoc2Z7aW59X0Z9IiwxXSxbMSwzLCJcXGxscHJ7Rn5cXG1hdGhzZntpbn1fRn0iLDFdXQ==
  \[\begin{tikzcd}
    {F A} && A \\
    \\
    {F (F A)} && {F A}
    \arrow["{\mathsf{in}_F}", from=1-1, to=1-3]
    \arrow["{F \llpr{F~\mathsf{in}_F}}"{description}, from=1-1, to=3-1]
    \arrow["{\llpr{F~\mathsf{in}_F}}"{description}, from=1-3, to=3-3]
    \arrow["{F~\mathsf{in}_F}", from=3-1, to=3-3]
  \end{tikzcd}\]
  So, substituting the second equation into the first, we have
  \[ \mathsf{in}_F \circ \llpr{F~\mathsf{in}_F} = \mathsf{id}_{F A}, \]
  completing the proof that $\mathsf{in}_F : F A \to A$ is an isomorphism.
\end{proof}

With Lambek's lemma in hand, we can now give a precise categorification of inductive types: the
inductive type represented by $\mu t.~\tau$ is interpreted as the initial algebra of the endofunctor
$F(t) = \llbr{\tau}$, where we extend our previous translation of types to map the free type
variable $t$ to the functor variable of the same name.\footnote{The action of $F$ on morphisms must
also be defined in the ``obvious'' way. We elide the details, but one can inspect the definition of
$F_\texttt{Nat}$ at the beginning of this subsection for an example.} The algebra's carrier $\mu F$
is the object representing the type, while its structure map $\mathsf{in}_F$ represents the type's
constructor.

It remains only to precisify our former account of recursors. Recall our previous observation that
recursors map out of an inductive type $\mu F$ into some other type $A$ using a ``combining
morphism'' that runs ``parallel'' to the constructor morphism---i.e., one that maps into $A$ from an
object that ``replaces'' instances of $\mu F$ with instances of $A$. Using our newfound language of
functors, we can phrase this much more concisely: a recursor maps from $\mu F$ to some type $A$
using a ``combining morphism'' $f_c : F A \to A$. Or, as a diagram:

% https://q.uiver.app/#q=WzAsNCxbMCwwLCJGKFxcbXUgRikiXSxbMiwwLCJcXG11IEYiXSxbMCwyLCJGIEEiXSxbMiwyLCJBIl0sWzAsMSwiYyJdLFsyLDMsImZfYyJdLFswLDIsIkYgKFxcbWF0aHNme2l0ZXJ9X3tmX2N9KSIsMV0sWzEsMywiXFxtYXRoc2Z7aXRlcn1fe2ZfY30iLDFdXQ==
\[\begin{tikzcd}
	{F(\mu F)} && {\mu F} \\
	\\
	{F A} && A
	\arrow["c", from=1-1, to=1-3]
	\arrow["{F (\texttt{rec}_{f_c})}"{description}, from=1-1, to=3-1]
	\arrow["{\texttt{rec}_{f_c}}"{description}, from=1-3, to=3-3]
	\arrow["{f_c}", from=3-1, to=3-3]
\end{tikzcd}\]

In fact, we already saw this diagram (diagrams (\ref{fig:bool-squares}) and (\ref{fig:nat-squares})
are variations of it in the particular cases of natural numbers and Booleans). But having now
developed the vocabulary of $F$-algebras, this should look familiar for another reason: recursors
are $F$-algebra morphisms! More precisely, given an inductive type $\mu F$, the ``unapplied''
recursor function is represented by $\llpr{-}_F$; given a particular ``combining morphism'' $f_c : F
A \to A$---which is precisely an $F$-algebra over $A$---the resulting parameterized recursor is
$\llpr{f_c} : \mu F \to A$.

As an example, we translate the parity function on natural numbers, which returns \texttt{true} if
its input is even and \texttt{false} if it is odd, as an application of the natural-number recursor.
First, we identify the appopriate ``combining function'' for this fold. Since we are ultimately
producing a Boolean, this function should translate to a morphism of type $F_\texttt{Nat}
\llbr{\texttt{Bool}} \to \llbr{\texttt{Bool}}$. We first define its programmatic version below:

\begin{align*}
\texttt{f}_\texttt{even} &: 1 \oplus \texttt{Bool} \to \texttt{Bool}\\
\texttt{f}_\texttt{even}~(\inl~()) &= \texttt{true}\\
\texttt{f}_\texttt{even}~(\inr~\texttt{b}) &= \texttt{not}(\texttt{b})
\end{align*}

This becomes the $F_\texttt{Nat}$-algebra
\[ \alpha_\texttt{even} = [\texttt{true}, \texttt{not}] \] which finally reveals our translation of
the parity function to be $\llpr{\alpha_\texttt{even}} : \llbr{\texttt{Nat}} \to
\llbr{\texttt{Bool}}$.

Note that \texttt{not} is, in turn, defined by an algebra over $F_\texttt{Bool}$, the constant-$1
\oplus 1$ functor. In particular, $\llbr{not} = \llpr{\alpha_\texttt{not}}$ where
$\alpha_\texttt{not} : 1 \oplus 1 \to 1 \oplus 1$ is $[\inr, \inl]$.

% \begin{exercise}
% \sloppy
%   Fill in the missing piece of the above translation by categorifying \texttt{not} using a suitable
%   algebra over the endofunctor characterizing Booleans.
% \end{exercise}

To put together the developments in this section, let's consider one more example of a categorified
inductive datatype, now expressed in terms of initial algebras. Recall the inductive datatype of
lists:

\begin{lstlisting}
inductive List (A : Type) : Type where
  | nil : List A
  | cons : A -> List A -> List A
\end{lstlisting}

Notice that this type is parameterized over a fixed type \texttt{A}; we will write $A$ for that
type's categorical denotation. We can read the endofunctor of which \texttt{List A} is a least fixed
point off of its constructors:\footnote{We again elide the action on morphisms.}
\[ F_{\texttt{List}_\texttt{A}}(X) = 1 \oplus A \times X. \]
Lists are thus represented by the initial algebra $(\mu F_{{\texttt{List}_\texttt{A}}},
\mathsf{in}_{F_{\texttt{List}_\texttt{A}}})$. Returning to our example from earlier, we can define the list
length function as the fold induced by the ``combining function''
\begin{align*}
  \texttt{f}_\texttt{length} &: \texttt{Unit} \oplus \texttt{A} \times \texttt{Nat} \to \texttt{Nat}\\
  \texttt{f}_\texttt{length}~(\inl~()) &= 0\\
  \texttt{f}_\texttt{length}~(\inr~(a, \ell)) &= \texttt{succ}(\ell)
\end{align*}
which we translate into the $F_{\texttt{List}_\texttt{A}}$-algebra
\[ \alpha_\texttt{length} = [z, s \circ \pi_1] \]
to obtain the fold $\llpr{\alpha_\texttt{length}} : \mu F_{\texttt{List}_\texttt{A}} \to \llbr{\texttt{Nat}}$.

\begin{exercise}
  \label{ex:self-trees}
Below, we define a strange type of so-called ``self-trees:'' each node can have as many children as
the parameterizing type has elements (i.e., a \texttt{SelfTree A} node can have, as it were,
``\texttt{A} many children'').
\begin{lstlisting}
inductive SelfTree (A : Type) : Type where
  | node : A -> (A -> SelfTree A) -> SelfTree A
  | leaf : SelfTree A
\end{lstlisting}

\begin{enumerate}
  \item Specify the functor $F_{\texttt{SelfTree}}$ that characterizes this type.
  \item We define the ``self-height'' of a self-tree to be the number of nodes encountered by
        successively visiting the $a$th child of a node containing $a$. (The self-height of a leaf
        is 0.) Categorify this operation using a suitable $F_{\texttt{SelfTree}}$-algebra.
\end{enumerate}
\end{exercise}

\begin{exercise}
\label{ex:app-as-algebra}
  Show that the morphism $\mathsf{app}$ induced by the universal property of exponentials is the
  structure map for an algebra whose carrier is the initial object. Use this to give a non-empty
  definition of an inductive type (i.e., an \texttt{inductive} declaration listing at least one
  putative constructor) that is isomorphic to the empty type.

  \emph{Hint:} Recall that in a CCC, $0^A \cong 0 \cong 0 \times A$.
\end{exercise}

\subsection{On the Existence of Initial Algebras}

Up to this point, we have simply assumed that, given an endofunctor $F$, we can always find an
initial $F$-algebra. This is not, however, always the case. And we should expect as much: after all,
not all recursive type equations admit a least fixed point---this is why we insist that no
non-positive occurrences of $t$ appear in $\tau$ when defining an inductive type $\mu t.~\tau$. The
failure to obtain a least fixed point in a recursive type equation corresponds to an analogous lack
of an initial algebra for the corresponding functor. For instance, $\mu t.~(t \to 2) \to 2$ (i.e.,
$F(X) = 2^{2^X}$) does not admit a least fixed point in every Cartesian closed category: for one,
Lambek's lemma says that this would induce an isomorphism $\mu F \cong 2^{2^{\mu F}}$, which by
Cantor's theorem never holds in the CCC $\Set$. In fact, we can more generally prove the
non-existence of certain initial algebras in arbitrary categories, not just $\Set$, but this would
take us too far afield. For the purposes of this chapter, it suffices to know that all the usual
strictly positive inductive types of interest are characterized by endofunctors that admit initial
algebras.

\begin{exercise}
\label{ex:isorec-translation}
\sloppy
There is another way to think about interpreting inductive types. While we gave an interpretation by
translating the various features of inductive types' structure into categorical terms, we could have
alternatively followed our previous method and directly translated the generalized inference rules
for isorecursively formulated inductive types (cf.\ Harper, \emph{Practical Foundations for
Programming Languages (2\textsuperscript{nd} ed.)}, Ch.\ 15).

Give a translation of these rules (shown below), and verify that it yields the same construction at
which we just arrived.

\begin{mathpar}
\inferrule*[right=T-fold]
  {\vdash t.\tau~\mathsf{pos}}
  {e : \tau[\mu t. \tau/t] \vdash \texttt{fold}(e) : \mu t.\tau}

\inferrule*[right=T-rec]
  {x : \tau[\rho/t] \vdash e_1 : \rho}
  {e_2 : \mu t.\tau \vdash \texttt{rec}(x. e_1)(e_2) : \rho}

\inferrule*[right=$\beta^{\texttt{rec}}$]
  {x : \tau[\rho/t] \vdash e_1 : \rho}
  {e_2 : \tau[\mu t. \tau/t] \vdash \texttt{rec}(x.e_1)(\texttt{fold}(e_2)) \equiv e_1[\texttt{map}(e_2)(y. \texttt{rec}(x.e_1)(y))/x] : \rho}

\inferrule*[right=$\eta^{\texttt{rec}}$]
  {y : \tau[\rho/t] \vdash e : \rho \\ e_2 : \tau[\mu t.\tau/t] \vdash e[\texttt{fold}(e_2)/y] \equiv e_1[\texttt{map}(e_2)(y. \texttt{rec}(x.e_1)(y))/x]}
  {y : \mu t.\tau \vdash e \equiv \texttt{rec}(x.e_1)(y) : \rho}
\end{mathpar}

Note that these rules introduce two new auxiliary definitions:
\gdef\NL{\\} % See https://tex.stackexchange.com/a/519322 and https://tex.stackexchange.com/a/234314
\begin{itemize}
  \item The judgment $\vdash t. \tau~\mathsf{pos}$ asserts that $t$ has only \emph{positive}
        occurrences in $\tau$; i.e., it never appears to the left of an arrow. Note that all of the
        rules implicitly carry this hypothesis, though we have only explicitly stated it for the
        first.
  \item The auxiliary function $\texttt{map}(e)(e')$ ``unpacks'' the structure of $e$ based on its
        type and applies $e'$ to each ``element'' thereof. Its typing rule is as follows (where
        $t$ may appear free in $\tau$):
        \begin{mathpar}
          \inferrule*[right=T-map]
            {\vdash e : \tau [\rho / t] \\ x : \rho \vdash e' : \rho'}
            {\vdash \texttt{map}(e)(x. e') : \tau[\rho'/t]}
        \end{mathpar}
        Here are two of its equational rules (it has one for every type former):
        \begin{mathpar}
          \inferrule*[right=$\beta^{\texttt{map}}_\times$]
            {\vdash e : (\tau_1 \times \tau_2)[\rho/t] \\ x : \rho \vdash e' : \rho'}
            {\vdash \texttt{map}(e)(e') \equiv \langle \texttt{map}(\mathsf{fst}~e)(x. e'), \texttt{map}(\mathsf{snd}~e)(x. e') \rangle \\: \tau_1[\rho'/t] \times \tau_2[\rho'/t]}
          
          \inferrule*[right=$\beta^{\texttt{map}}_\oplus$]
            {\vdash e : (\tau_1 \oplus \tau_2)[\rho/t] \\ x : \rho \vdash e' : \rho'}
            {
              \begin{aligned}
                \vdash \texttt{map}(e)(e') \equiv~ &\texttt{case}~e~\texttt{of}\NL[-6pt]
                &|~\texttt{inl}~x_1 \mapsto \texttt{inl}~(\texttt{map}(x_1)(x. e'))\NL[-6pt]
                &|~\texttt{inr}~x_2 \mapsto \texttt{inr}~(\texttt{map}(x_2)(x. e')) \NL[-6pt]
                &: \tau_1[\rho'/t] \oplus \tau_2[\rho'/t]
              \end{aligned}
            }
        \end{mathpar}
        % Note that we are slightly abusing notation here: the second argument of \texttt{map} binds a
        % variable $x$, which we have elided in syntax. This is intended to be more suggestive of the
        % categorical translation.
\end{itemize}
Your translation should give an account of what these definitions mean categorically.

\vspace{12pt}

\noindent Here are some hints for this exercise:
\begin{itemize}
  \item If parameterized types are translated as functors, to what does type-variable substitution
        correspond?
  \item $\texttt{map}(e)$ will not be translated as a morphism.
\end{itemize}
\end{exercise}

% \begin{exercise}
% \sloppy
% (Note that this exercise is independent of the preceding one.)

% \noindent
%   We now face the perennial categorical question: can we dualize what we have done? (Spoiler: yes.)
%   That is, given an endofunctor $F$, let us consider \emph{$F$-coalgebras} $(A,\, \eta : A \to F A)$
%   instead of $F$-algebras, and examine the \emph{terminal} $F$-coalgebra rather than the initial
%   one.

%   \begin{enumerate}
%     \item What type does the carrier of the terminal $F$-coalgebra represent? (\emph{Hint}: It will
%           turn out that the carrier of the terminal $F$-coalgebra is also a fixed point of $F$ up to
%           isomorphism.) What about its structure map?

%     \item How do we interpret the unique $F$-coalgebra morphism from another $F$-coalgebra into the
%           terminal one?
%   \end{enumerate}
% \end{exercise}

\section{The Families Fibration: Indexing Inductive Types}

Having given a categorical translation of simple inductive types, we now turn our attention to
\emph{indexed} inductive types. Recall that indexed inductive types enrich inductive types by
allowing types to be parameterized by data values at the type level: for instance, in addition to
the simple type \texttt{List A} of arbitrary-length lists with elements of type \texttt{A}, we can
also define the indexed type \texttt{Vector A n} of \texttt{A}-containing lists of length
\texttt{n}.

% This is the only feature of a dependent type system we will require, so we give a categorical
% account specifically of indexed inductive types rather than full dependent types. The advantage of
% this simplification is that such an account follows naturally from our preceding work.

While we gave a generalized presentation of inductive types in terms of initial algebras of
endofunctors over arbitrary categories, hereafter we will specifically consider a categorical
semantics in terms of sets. While all of the ideas presented can be generalized to other categories,
focusing on $\Set$ will simplify some of the required categorical machinery. (Note that our
preceding interpretation of inductive types can indeed be carried out in $\Set$, since it is
Cartesian closed and contains finite coproducts---i.e., it is \emph{bicartesian closed}.)

The category $\Set$ alone, however, lacks the structure to give an elegant semantics to indexed
inductive types. Instead, we will turn to a new category built atop $\Set$ that is equipped with a
notion of ``indexing'' that will afford a straightforward interpretation of our inductive types'
indices.\footnote{Be careful when reading the following definition and elsewhere in this chapter:
the notation $\prod_{a \in A} S(a)$ below refers to a \emph{set-theoretic} indexed product, not,
e.g., a $\Pi$-type. The symbols $\prod$, $\sum$, and $\to$ are regrettably highly overloaded (we
will see yet another denotation of $\sum$ later on), so the meanings of these notations are highly
context-sensitive.}

\begin{definition}[Families fibration]
  The \emph{category of families (of sets)} $\FamSet$ is defined as follows:
  \begin{itemize}
    \item Objects are pairs $(A, P : A \to \Ob(\Set))$ of an indexing set $A$ and a \emph{family} of
          sets $P$ specifying a set $P(a)$ for each index $a \in A$.
    \item Morphisms from $(A, P)$ to $(B, Q)$ are pairs of functions $(f, f^\sim)$ such that $f : A
          \to B$ and $f^\sim : \prod_{a \in A} P(a) \to Q(f(a))$.
  \end{itemize}

  \noindent The forgetful functor $U : \FamSet \to \Set$ defined by
  \begin{align*}
    U(A, P) &= A\\
    U(f, f^\sim) &= f
  \end{align*}
  is called the \emph{families fibration}.
\end{definition}

Note that, while we will not treat them in full generality, \emph{(Grothendieck) fibrations} are
objects of particular interest in category theory and characterize a certain kind of structure
(which we will attempt to intuitively motivate at several points hereafter). Throughout this
chapter, you will encounter definitions phrased as features ``of the families fibration:'' these
should be read as referring to the entire structure of $\FamSet$, $\Set$, and this mapping between
them, rather than particular characteristics of a single forgetful functor.

% \begin{definition}[Families fibration (of $\Set$)]
%   The \emph{families fibration (of $\Set$)} is a functor $U : \FamSet \to \Set$ defined as follows:
%   \begin{align*}
%     U(A, P) &= A\\
%     U(f, f^\sim) &= f
%   \end{align*}
% \end{definition}

Let's now use the category of families to interpret indexed inductive types. We'll start with the type
\texttt{BitVec n}, which is like the type \texttt{Vector A n} we have previously discussed, except
that it contains only Booleans:\footnote{\texttt{Vector A n} is defined similarly. We choose
\texttt{BitVec} for this example to avoid any potential confusion between the parameter \texttt{A}
and the index \texttt{n} of \texttt{Vector}.}

\begin{lstlisting}
inductive BitVec : Nat -> Type
  | bvnil  : BitVec 0
  | bvcons : {n : Nat} -> Bool -> BitVec n -> BitVec (succ n)
\end{lstlisting}

Drawing inspiration from our preceding development of simple inductive types, we might expect to
interpret this type as an endofunctor on $\FamSet$. That endofunctor should produce a family that
looks like a representation of the Boolean list type (i.e, the sum of an ``empty'' object and a
Boolean--vector pair object), except that we can only inject into the left-hand side of the sum at
index zero and into the right-hand side at successor indices. Concretely, such a functor looks as
follows:\footnote{We will continue our convention of eliding morphism actions. As before, these
mappings are ``automatic'' from our choice of action on objects.}

\begin{align}
  F_\texttt{BitVec}(\N, X) = (\N, (n \mapsto
    &\left(\begin{array}{ll}\{\star\} & \text{if } n = 0 \\ \varnothing & \text{else} \end{array}\right)
    \oplus \notag\\
    &\left(\begin{array}{ll} \varnothing & \text{if } n = 0 \\ \{n_1\} \times \llbr{\texttt{Bool}} \times X(n_1) & \text{if } n = \texttt{succ}(n_1) \end{array} \right)
  ))
  \label{eqn:functor-bitvec}
\end{align}

But here we encounter a problem: the endofunctor we have just defined only acts on $\N$-indexed
families. Therefore, it is \emph{not} an endofunctor on $\FamSet$, whose objects can have arbitrary
indexing sets. Yet this $\N$-indexing really is essential: our functor cannot perform a case split
on zero and successors if our indexing set does not have such elements! Indeed, this ability to
analyze the structure of a type index is precisely one of the key attributes that makes indexed
inductive types more powerful than traditional parametric polymorphism.

Therefore, to represent an indexed inductive type, we take the initial algebra of an endofunctor not
on $\FamSet$ itself, but on a restriction of $\FamSet$ where we consider only families indexed by
the indexing set of the indexed type of interest. This smaller category is called the \emph{fiber}
of the families fibration over $A$.

\begin{definition}[Fiber of the families fibration over $A$]
\sloppy
  The \emph{fiber} of the families fibration over $A$, written $\FamSet_A$, is a subcategory of
  $\FamSet$ in which all objects have indexing set $A$. Precisely:
  \begin{itemize}
    \item Objects are families $(A, P : A \to \Ob(\Set))$
    \item Morphisms are pairs $(\id_A, f^\sim : \prod_{a \in A} P(a) \to Q(a))$
  \end{itemize}
\end{definition}

Thus, although it will at times be useful to consider the category $\FamSet$ in its entirety, we
will represent $A$-indexed inductive types by initial algebras of endofunctors on just the fiber of
the families fibration over $A$. As before, the morphisms induced by initiality correspond to folds
using the target algebras' structure maps.

\begin{exercise}
\label{ex:fin-functor}
\sloppy
Identify the endofunctor on $\FamSet_\N$ that characterizes \texttt{Fin n}, the type of ``natural
numbers less than $n$'' (or, isomorphically, $\Z/n\Z$):
\begin{lstlisting}
inductive Fin : Nat -> Type where
  | finZero : {n : Nat} -> Fin (succ n)
  | finSucc : {n : Nat} -> Fin n -> Fin (succ n)
\end{lstlisting}

\emph{Hint:} Notice that, for a given index, there maybe multiple (or no) applicable constructors.
\end{exercise}

Notice that we can also interpret refinements of simple types in the category of families. If $T =
\mu F$ is an inductive type, then given an $A$-valued combining function $f$ for $T$ (i.e., an
$F$-algebra $f$ with carrier $A$) and a value $a : A$, the refinement of $T$ where $f$ yields value
$a$, i.e.,
\[ \texttt{\{t : T | }\texttt{rec}_\texttt{A}\texttt{(f)(t) = a} \texttt{\}} \]
can be translated as the family
\[ (A, a \mapsto \{ x \in \mu F \mid \llpr{f}_{F}(x) = a \}). \]
This is the ``na\"ive'' subset interpretation of refinements that we discussed previously. Our
objective will be to turn this family into one that is the denotation of an indexed inductive type.

\section{Constructing a Refinement Translation}

Our basic recipe for the translation of $\texttt{\{t : T | }\texttt{rec}_\texttt{A}\texttt{(f)(t) =
a} \texttt{\}}$ into an indexed inductive type will be as follows:
\begin{enumerate}
  \item Lift $F$, an endofunctor on $\Set$, to an endofunctor $\hat F_A$ on $\FamSet_A$.
  \item Restrict the family produced by our lifting to those elements which are mapped to the
        appropriate value by $\texttt{rec}_\texttt{A}\texttt{(f)}$.
\end{enumerate}
This recipe will require a number of auxiliary constructions, on which we now embark.

\subsection{Another View of Families}

For some of the constructions we give in this chapter, it will be helpful to have in mind an
alternative perspective on $\FamSet$. This perspective is inspired by the Curry--Howard
correspondence between propositions and types.

Recall that the Curry--Howard correspondence suggests another way to interpret indexed inductive
types: as \emph{predicates} on the indexing type. For instance, we can define what it means for a
natural number to be even by declaring a type of \emph{proofs} of evenness:

\begin{lstlisting}
inductive Even : Nat -> Type where
  | evZero : Even 0
  | evSucc : {n : Nat} -> Even n -> Even (succ (succ n))
\end{lstlisting}

A natural number \texttt{n} is even if the type \texttt{Even n} is inhabited. Correpsondingly, the
interpretation of this inductive type as a family of sets $(\N, P_\texttt{Even})$ can be thought of
as defining a \emph{predicate} $P_\texttt{Even}$ on natural numbers. With this interpretation, we
think of a family $(A, P)$ as containing proofs $P(a)$ that the represented predicate holds of
elements $a \in A$. In particular, if $P(a)$ is inhabited, than the predicate holds of $a$; if it is
empty, the predicate does not.

Adopting the logical perspective on families suggested by the Curry--Howard correspondence, it is
natural to consider the family that represents the constantly-true predicate on a given set, known
as the set's \emph{truth predicate}.
\begin{definition}[Truth predicate and truth functor]
\sloppy
  Given a set $A$, the \emph{truth predicate} for $A$ is the family $(A, a \mapsto 1)$, where 1 is
  the terminal object in $\Set$ (i.e., $\{\star\}$). The truth predicate is terminal in $\FamSet_A$.

  The functor $\top : \Set \to \FamSet$ that sends sets to their truth objects (and maps morphisms
  by pairing them with $\_ \mapsto \mathsf{id}_1$) is called the \emph{truth functor} for the
  families fibration.
\end{definition}

\subsection{Comprehension and Reindexing}

Returning to our ``recipe'' from earlier, our first order of business is, given an endofunctor $F :
\Set \to \Set$ with an initial algebra, to define an endofunctor $\hat F_A : \FamSet_A \to
\FamSet_A$ whose initial algebra has a carrier representing an $A$-indexed type that ``looks like''
the inductive type characterized by $F$.

To define a translation between $\Set$ and $\FamSet$, we need to be able to move between these two
categories. We have already seen one functor in each direction: the truth predicate $\top : \Set \to
\FamSet$ maps sets to the trivial family thereover, while the families fibration $U : \FamSet \to
\Set$ discards the family associated with an indexing set. We will now introduce one additional
functor in the backward direction, which exploits the untypedness of $\Set$ to ``collapse'' a family
in $\FamSet$ into a single object in $\Set$.

\begin{definition}[Comprehension functor]
\sloppy
  The \emph{comprehension} of a family $(A,P)$, written $\{(A, P)\}$, is the set $\sum_{a \in A}
  P(a)$. We correspondingly define the \emph{comprehension functor} $\{-\} : \FamSet \to \Set$,
  which sends a family to its comprehension and a pair of morphisms to a morphism on pairs.
\end{definition}

The comprehension functor $\{-\}$ and forgetful functor $U$ are related by the natural
transformation $\pi : \{-\} \Rightarrow U$ whose morphisms project (hence the notation $\pi$) the
index from an element of a (``comprehended'') indexed family:
\[ \pi_{(A,P)}(a, p) = a \]

We will also require means of mapping between $\FamSet_A$ and $\FamSet_B$ for distinct indexing sets
$A$ and $B$. Intuitively, this is because we interpret our indexed inductive types as endofunctors
only over a specific fiber, so we will need to move between fibers in order to, for instance, fold
over an indexed inductive type into a distinct output type. This will be accomplished using another
functor, known as the \emph{op-reindexing functor}.

\begin{definition}[Indexed coproduct]
% \sloppy % Not worth it
  Let $f : A \to B$ be a set function, and let $(A, P)$ be a family of sets. The \emph{indexed
  coproduct of $(A, P)$ along $f$}, written $\Sigma_f (A, P)$, is defined as
  \[ \left(B, b \mapsto \sum_{a \in A} (b = f(a)) \times P(a)\right) \]
  where $(b = f(a))$ is a set that is inhabited if the equality holds and uninhabited if not.

  This definition induces the \emph{op-reindexing functor} $\sum_f : \FamSet_A \to \FamSet_B$, whose
  action on morphisms lifts functions to act on the relevant component of tuples:
  \[ \Sigma_f (\id_A, g^\sim) = (\id_B, b \mapsto (a, h, p) \mapsto (a, h, g^\sim(a)(p))) \]
\end{definition}

In other words, $\Sigma_f (A, P)$ maps a given $b$ in the image of $f$ to the members of the family
$P$ at each element of $b$'s preimage. Indeed, an isomorphic (and perhaps more intuitive)
characterization of $\Sigma_f (A, P)$ is given by
\begin{equation}
  \label{eqn:indexed-coprod-simplified}
  \left(B, b \mapsto \sum_{a \in f^{-1}[\{b\}]} P(a)\right).
\end{equation}

We summarize the developments of this section in the diagram below: the category of families sits
``above'' the category of sets, with each fiber situated atop its corresponding index set. As
discussed previously, we have functors running in ``orthogonal'' directions, with $U$, $\top$, and
$\{-\}$ mapping between $\FamSet$ and $\Set$ while $\Sigma_f$ reindexes between fibers. We will
have more to say about this diagram later.

\begin{center}
\begin{tikzpicture}
  \node (FamSet) at (0, 1.75) {};
  \node (Set) at (0, -1.5) {};
	\draw[draw=black, thin, solid] (FamSet) ellipse (3.75 and 1.75);
	\draw[draw=black, thin, solid] (Set) ellipse (1.50 and 0.50);
	\node[black, anchor=south west] at (1.3,-1.3) {$\Set$};
	\node[black, anchor=south west] at (1.5,3.25) {$\FamSet$};
	\node[black, anchor=south] at (-1.75,2.4) {$\FamSet_A$};
	\node[black, anchor=south] at (1.75,2.4) {$\FamSet_B$};
  \node (FamSetA) at (-1.75, 1.5) {};
  \node (FamSetB) at (1.75, 1.5) {};
	\draw[draw=black, thin, solid] (FamSetA) ellipse (1 and 1);
	\draw[draw=black, thin, solid] (FamSetB) ellipse (1 and 1);
	\draw[draw=black, thin, dashed] (-0.5,-1.3) -- (-1.5,0.50);
	\draw[draw=black, thin, dashed] (0.5,-1.3) -- (1.5,0.50);
	\node[black] (A) at (-0.5,-1.5) {$A$};
	\node[black] (B) at (0.5,-1.5) {$B$};
	\node[black] at (-1.75,1.75) {$(A, P)$};
	\node[black] at (-1.75,1.25) {$(A, P')$};
	\node[black] at (1.75,1.75) {$(B, Q)$};
	\node[black] at (1.75,1.25) {$(B, Q')$};
  \draw[draw=black, ->] (A) -- (B) node [midway, below=2pt, label=$f$] {};
  \draw[draw=black, ->] (FamSetA)++(1,0) -- (0.75, 1.5) node [midway, below=2pt, label=$\Sigma_f$] {}; 
  \draw[draw=black, <-] (Set)++(-1,0.5) -- (-1, 0) node [midway, below left=6pt, label=$U$] {};
  \draw[draw=black, ->] (Set)++(0,0.5) -- (0, 0) node [midway, below left=6pt, label=$\top$] {};
  \draw[draw=black, <-] (Set)++(1,0.5) -- (1, 0) node [midway, below right=10pt, label=$\{-\}$] {};
  % Hacky nat trans arrow b/c we don't have tikzcd
  \draw[draw=black] (-0.85, -0.8) -- (0.9, -0.8) node [midway, below right=5pt, label=$\pi$] {};
  \draw[draw=black] (-0.85, -0.86) -- (0.9, -0.86);
  \node[black] at (-0.8, -0.85) {\footnotesize $\Leftarrow$};
\end{tikzpicture}
\captionof{figure}{A summary of functors to be used in our translation.}
\label{fig:famset-diag-no-adjoints}
\end{center}

\begin{exercise}
  \label{ex:very-strong-coprods}
  Fix a family $(A, P)$. Let $f : A \to B$ be a set function. Consider the following diagram, where
  $\psi : (A, P) \to \Sigma_f (A, P)$ acts in the straightforward way:

  % https://q.uiver.app/#q=WzAsNCxbMCwwLCJcXHsoQSxQKVxcfSJdLFsyLDAsIlxce1xcU2lnbWFfZihBLCBQKVxcfSJdLFswLDIsIkEiXSxbMiwyLCJCIl0sWzAsMSwiXFx7XFxwc2lcXH0iXSxbMCwyLCJcXHBpX3soQSwgUCl9IiwyXSxbMiwzLCJmIiwyXSxbMSwzLCJcXHBpX3tcXFNpZ21hX2YoQSwgUCl9Il1d
  \[\begin{tikzcd}
    {\{(A,P)\}} && {\{\Sigma_f(A, P)\}} \\
    \\
    A && B
    \arrow["{\{\psi\}}", from=1-1, to=1-3]
    \arrow["{\pi_{(A, P)}}"', from=1-1, to=3-1]
    \arrow["{\pi_{\Sigma_f(A, P)}}", from=1-3, to=3-3]
    \arrow["f"', from=3-1, to=3-3]
  \end{tikzcd}\]

  \begin{enumerate}
    \item Show that this diagram commutes.
    \item Show that $\{\psi\}$ is an isomorphism.
  \end{enumerate}

  This shows that the families fibration has \emph{very strong coproducts}.
\end{exercise}

\subsection{Liftings}

We now have all the tools required to define a lifting of an endofunctor $F$ on $\Set$ to an
endofunctor $\hat F$ on $\FamSet$, which we may then restrict to $\FamSet_A$. (It makes sense that
we can lift ``all the way'' to $\FamSet$, rather than just a particular fiber $\FamSet_A$, since the
type whose endofunctor we are lifting does not depend on a particular index.) Before we attempt to
define such a lifting, however, we should first set out the desiderata we expect such a construction
to satisfy.

As a basic criterion, we ought to expect that the indexing set produced by our ``lifted''
endofunctor $\hat F$ is the set obtained by simply applying $F$ to the original indexing set.
Indeed, we can now drop the scare quotes: this is precisely the definition of the lifting of an
endofunctor to $\FamSet$.

\begin{definition}[Lifting]
  A \emph{lifting} of an endofunctor $F : \Set \to \Set$ is an endofunctor $\hat F : \FamSet \to
  \FamSet$ such that $F \circ U = U \circ \hat F$.
\end{definition}

However, this criterion says nothing about the behavior of the \emph{family} produced by $\hat F$
itself. Adopting the logical view again, we can formulate one such criterion: $\hat F$ should
``respect truth.'' That is, if we think of $F(A)$ (given some $A$) as implicitly defining a
predicate (namely, one which holds of all its elements), then $\hat F$ should yield the same
predicate---i.e., it should yield a predicate that holds of the same values---when ``applied to''
$A$. Of course, this intuitive phrasing does not type-check: $F(A)$ is not a predicate (it is a
set), and $\hat F$ cannot be applied to $A$ (since $A$ is not a family). But we can fill in some
straightforward coercions to make this definition---which characterizes a so-called
\emph{truth-preserving} lifting---precise: to treat the objects in the codomain of $F$ as
predicates, we post-compose the truth functor; to allow $\hat F$ to act on sets, we pre-compose the
truth functor; and the notion that these functors always yield the ``same predicate'' as each other
is captured by natural isomorphism.

\begin{definition}[Truth-preserving lifting]
  Let $F : \Set \to \Set$ be an endofunctor. A lifting of $F$ is \emph{truth-preserving} if there
  exists a natural isomorphism $\top \circ F \cong \hat F \circ \top$.
\end{definition}

\begin{exercise}
\label{ex:trivial-truth-pres}
  Verify that
  \begin{align*}
    \hat F(A, P) &= (F A,\, \_ \mapsto 1)\\
    \hat F(f, f^\sim) &= (f,\, \_ \mapsto \id_1)
  \end{align*}
  defines a trivial truth-preserving lifting of $F : \Set \to \Set$.
\end{exercise}

The truth-preserving lifting above, however, is degenerate in the sense that it entirely forgets the
structure of the original family. We therefore instead define the following truth-preserving
lifting:\footnote{One can find a description of the lifting's action on morphisms, along with a proof
that it is truth-preserving, in~\citep{ghani2010fibrational}.}

% We define a truth-preserving lifting of an arbitrary endofunctor $F : \Set \to \Set$ as
% follows:\footnote{We elide its action on morphisms and the proof that it is truth-preserving; both
% can be found in Ghani et al.~\citep{ghani2010fibrational}.}

\[ \hat F(A, P) = \Sigma_{F\pi_{(A, P)}} \top(F\{(A, P)\}) \]

Unlike in the exercise, we incorporate the ``structure of $P$ under $F$'' by applying $F$ to the
\emph{comprehension} of the original family. We then re-separate the indexing set $F A$ and
associated family ``under $F$'' by taking an indexed coproduct along $F \pi_{(A, P)}$. Thus, the
resulting family can be seen as containing those elements which ``wrap the structure of $P$ in $F$''
(or, from a logical perspective, the resulting predicate holds only of those elements of $F A$ whose
``contents'' in $A$ were validated by $P$). Unfolding definitions makes these intuitions more
apparent:

\begin{align*}
  \hat F(A, P) &= \Sigma_{F\pi_{(A, P)}} \top(F\{(A, P)\})\\
    &= \Sigma_{F\pi_{(A, P)}} \left(F\{(A, P)\}, \_ \mapsto 1\right)\\
    &= \left(F A, x \mapsto \sum_{y \in F\{(A, P)\}} (F\pi_{(A, P)}(y) = x) \times 1 \right)\\
    &\cong (FA, x \mapsto \{y \in F\{(A, P)\} \mid (F \pi_{(A,P)})(y) = x\})
\end{align*}

As a concrete example, consider the truth-preserving lifting of $F_\texttt{Nat}$:

\begin{align*}
  \hat F_\texttt{Nat}(A, P) &= \Sigma_{F_\texttt{Nat}} \top(\hat F_\texttt{Nat}\{(A, P)\})\\
    &= \Sigma_{F_\texttt{Nat} \pi_{(A, P)}} \left(1 \oplus \{(A, P)\},\, \_ \mapsto 1\right)\\
    &\cong \left(1 \oplus A,\, b \mapsto \sum_{a \in F_\texttt{Nat} \pi_{(A, P)}^{-1}[\{b\}]} 1 \right)\\
    &\cong \left(1 \oplus A, \begin{cases}\inl~() \mapsto \inl~()\\ \inr~a \mapsto \{\inr~((a, p), \star) \mid p \in P(a)\} \end{cases}\right)\\
    &\cong \left(1 \oplus A, b \mapsto \left(\begin{array}{ll}1 & \text{if } b = \inl~() \\ \varnothing & \text{if } b = \inr~a \end{array}\right)
           \oplus \left(\begin{array}{ll} \varnothing & \text{if } b = \inl~() \\ \{a\} \times P(a) & \text{if } b = \inr~a \end{array}\right)\right)
\end{align*}

Finally, since our objective was to obtain an endofunctor on $\FamSet_A$---not $\FamSet$---we define
the functor $\hat F_A$ to be the restriction of $\hat F$ to $\FamSet_A$.

\subsection{Translating Refinements}

Recalling our original objective---to translate type refinements into indexed inductive types---we
can see that truth-preserving liftings get us halfway to our goal: they transform the endofunctors
that characterize simple inductive types into analogous endofunctors on families. But our aim is
not to produce an indexed inductive type that corresponds to the unrestricted simple one, but rather
some refinement thereof. This refinement translation is the last piece we must develop.

In fact, we already have the tool we need to do this: op-reindexing. Recall that indexed coproducts
(or, more generally, the op-reindexing functor) let us take a family and reindex it, such that the
resulting family contains only those elements which are mapped by the reindexing function to a given
value in its codomain. This is precisely what we do when we refine a datatype: given $\texttt{P} :
\texttt{T} \to \texttt{A}$, the type $\{ \texttt{t} : \texttt{T} \mid \texttt{P(t) = a} \}$ is an
$\texttt{A}$-indexed type that, at a given index $\texttt{a}$, consists of the elements of
$\texttt{T}$ on which $\texttt{P}$ produces value $\texttt{a}$. In other words, op-reindexing by $f$
corresponds precisely to refinement by $f$.

Therefore, the construction to which we have building all this time can finally be written as
follows:

\begin{definition}[Refinement translation]
  Let $F : \Set \to \Set$ be an endofunctor characterizing the inductive type $\mu F$. Let $(A,
  \alpha : F A \to A)$ be an $F$-algebra. We define the translation of the refinement type
  characterized by the family
  \[ (A, a \mapsto \{ x \in \mu F \mid \llpr{\alpha}_F(x) = a \}) \] as the indexed inductive type
  $\mu F^\alpha$ characterized by the endofunctor $F^\alpha : \FamSet_A \to \FamSet_A$ such that
  \[ F^\alpha = \Sigma_\alpha \circ \hat F_A. \]
\end{definition}

Once again, it is illuminating to expand this functor's action on objects:

\begin{align}
  F^\alpha(A, P) &= (\Sigma_A \circ \hat F_A)(A, P) \notag\\
    &\cong \Sigma_\alpha (FA, x \mapsto \{y \in F\{(A, P)\} \mid (F \pi_{(A,P)})(y) = x\}) \tag{As before} \notag\\
    &= \left(A, a \mapsto \sum_{x \in FA} (a = \alpha(x)) \times \{y \in F\{(A, P)\} \mid (F \pi_{(A,P)})(y) = x\}\right) \notag\\
    &\cong (A, a \mapsto \{ y \in F\{(A, P)\} \mid \alpha((F \pi_{(A, P)})(y)) = a \})
    \label{eqn:translation-expanded}
\end{align}

Upon cursory inspection, at least, the general shape of this family should seem reasonable: it is
$A$-indexed, with elements drawn from a set of ``$F$-shaped objects'' restricted to those ``on
which'' (after a certain projection) $\alpha$ attains value $a$.

We illustrate how this works in practice with our example from the beginning: deriving the type of
length-indexed vectors from the refinement of lists by their length. To simplify, we will consider
just lists of Booleans, and attempt to derive the type \texttt{BitVec} from before.

First, letting $2 \cong 1 \oplus 1$ be the object denoted by the Boolean type, we recall the
definition of $F_\texttt{BitList} : \Set \to \Set$ and the algebra $(\N, \texttt{length})$
thereover:
\begin{align*}
  F_\texttt{BitList}(X) &= 1 \oplus 2 \times X\\
  F_\texttt{BitList}(f) &= [\inl \circ \id_1, \inr \circ \langle \mathsf{id}_2, f\rangle]\\
  \\[-10pt] % Spacing hack
  % \texttt{length} : F_\texttt{BitList} \N \to \N &= [\texttt{z}, \texttt{s} \circ \pi_2]
  \texttt{length} ~ (\inl~ ()) &= 0\\
  \texttt{length} ~ (\inr~ (b, \ell)) &= \texttt{succ}(\ell)
\end{align*}

We now compute the corresponding indexed inductive type. Note that we use the isomorphic expansion
from equation (\ref{eqn:translation-expanded}) to avoid re-unfolding definitions:

\begin{align*}
  &F^\texttt{length}_\texttt{BitList}(\N, P)\\
  &\cong (\N, n \mapsto \{ y \in F_\texttt{BitList}\{(\N, P)\} \mid \texttt{length}((F_\texttt{BitList} \pi_{(\N, P)})(y)) = n\})\\
  &= (\N, n \mapsto \left\{ y \in 1 \oplus 2 \times (\N, P) \mid (y = \inl~() \text{ and } 0 = n) \text{ or } (y = \inr~(b, (n_1, p)) \text{ and } \texttt{succ}(n_1) = n) \right\})\\
  &= \left(\N, \begin{cases}0 \mapsto \{\inl~()\} \\ \mathsf{succ}(n_1) \mapsto \{\inr~(b, (n_1, p)) \mid b \in 2, p \in P(n_1)\} \end{cases} \right)\\
  &= (\N, (n \mapsto
    \left(\begin{array}{ll}1 & \text{if } n = 0 \\ \varnothing & \text{else} \end{array}\right)
    \oplus\\
    &\hspace{5.75em}\left(\begin{array}{ll} \varnothing & \text{if } n = 0 \\ 2 \times (\{n_1\} \times P(n_1)) & \text{if } n = \texttt{succ}(n_1) \end{array} \right)
  ))
\end{align*}

Up to isomorphism (specifically, with a minor rearrangement and reassociation in the right-hand
summand), this is precisely the functor $F_{\texttt{BitVec}}$ derived in equation
(\ref{eqn:functor-bitvec})!

\begin{exercise}
\label{ex:bh-indexed-rbtrees}
\sloppy
  Consider the following inductive type of trees with colored nodes (we refrain from calling these
  ``red-black trees'' in prose because they need not maintain the relevant invariants):
\begin{lstlisting}
  inductive RBList (A : Type) where
    | leaf  : RBTree
    | red   : A -> RBTree -> RBTree -> RBTree
    | black : A -> RBTree -> RBTree -> RBTree
\end{lstlisting}

  \begin{enumerate}
    \item Define the endofunctor $F_\texttt{RBTree} : \Set \to \Set$ such that
          $\llbr{\texttt{RBTree}} \cong \mu F_\texttt{RBTree}$.
    \item Define an $F_\texttt{RBTree}$-algebra $\texttt{bh}$ that computes the \emph{black height}
          of an \texttt{RBTree}, i.e., the maximum number of black nodes on a path from the root to
          a leaf. You may assume the existence of a function $\texttt{max} : \N \times \N \to \N$.
    \item Using the above, compute $F_\texttt{RBTree}^\texttt{bh} : \FamSet_\N \to \FamSet_\N$, and
          read off of this expression an indexed inductive type of black height-indexed colored
          trees. Verify that your definition matches how you would intuitively define this type from
          scratch.
  \end{enumerate}
\end{exercise}

\section{Correctness}

These illustrative examples are compelling, but we have yet to prove that our foregoing construction
always characterizes the indexed inductive type of interest. To begin with, we do not even know that
the functor $F^\alpha$ admits an initial algebra in the first place, nor---if it does---whether the
carrier of this algebra is in fact isomorphic to our na\"ive subset formulation of the refinement of
$\mu F$ by $\alpha$. In this section, we will sketch the proof of these facts, referring the reader
to the paper by Atkey et al.~\citep{atkey2011refining} for further details.

\subsection{Preliminaries}

We begin with a few auxiliary lemmas and definitions we will require for our proofs. First, recall
our observation from the previous section that the functors $\top : \Set \to \FamSet$ and $U :
\FamSet \to \Set$ map in oppposite directions. In fact, we can say more: these form an adjoint pair.

\begin{lemma}
  \label{lem:fam-fibration-adjoint}
  The functors $U \dashv \top$ form an adjoint pair.
\end{lemma}

\begin{proof}
  We must show $\Set(U(A, P), B) \cong \FamSet((A, P), \top B)$, i.e.,
  \[ \Set(A, B) \cong \FamSet((A, P), (B, \_ \mapsto 1)).\]
  Elements of the right-hand set are pairs of morphisms $(f : A
  \to B, f^\sim : \prod_{a \in A}, P(a) \to 1)$. But by terminality, there is only ever one choice
  of $f^\sim$, so the right-hand side is isomorphic to the set of functions $f : A \to B$, i.e.,
  $\Set(A, B)$. Naturality is straightforward.
\end{proof}

As before, we now turn from functors between $\FamSet$ and $\Set$---i.e., between families and their
index sets---to those between fibers $\FamSet_A$ and $\FamSet_B$---i.e., between families with
distinct index sets. While our construction in the previous section required only the op-reindexing
functor $\Sigma_f$, the following alternative means of reindexing will be useful for our proof.

\begin{definition}[Reindexing functor]
  Let $f : A \to B$ be a function (of sets). The \emph{reindexing functor} $f^* : \FamSet_B \to
  \FamSet_A$ is defined by
  \begin{align*}
    f^*(B, Q) &= (A, Q \circ f)\\
    f^*(\id_B, g^\sim) &= (\id_A, g^\sim \circ f)
  \end{align*}
\end{definition}

Notice that this functor, in contrast with $\Sigma_f$, is contravariant: the domain of $f^*$ is the
fiber over the codomain of $f$, and vice versa. (Hence the ``op'' prefix in ``op-reindexing.'')
Thus, given a function $f : A \to B$, we can produce functors running in both directions between
$\FamSet_B$ and $\FamSet_A$. Moreover, these functors are adjoint.

\begin{lemma}
  \label{lem:reindex-adjoint}
  Given a function $f : A \to B$, the functors $\Sigma_f \dashv f^*$ form an adjoint pair.
%   % https://q.uiver.app/#q=WzAsMyxbMCwwLCJcXEZhbVNldF9CIl0sWzIsMCwiXFxGYW1TZXRfQSJdLFsxLDAsIlxccGVycCJdLFsxLDAsIlxcU2lnbWFfZiIsMix7ImN1cnZlIjozfV0sWzAsMSwiZl4qIiwyLHsiY3VydmUiOjN9XV0=
% \[\begin{tikzcd}
% 	{\FamSet_B} & \perp & {\FamSet_A}
% 	\arrow["{f^*}"', curve={height=18pt}, from=1-1, to=1-3]
% 	\arrow["{\Sigma_f}"', curve={height=18pt}, from=1-3, to=1-1]
% \end{tikzcd}\]
\end{lemma}
\begin{proof}
  We show
  \[ \FamSet_B(\Sigma_f(A, P), (B, Q)) \cong \FamSet_A((A, P), f^*(B, Q)). \]
  Recall that morphisms in these fibers are fully characterized by their second component (i.e.,
  their mapping on families), so, unfolding definitions (and using the isomorphic representation of
  the indexed coproduct from equation (\ref{eqn:indexed-coprod-simplified})), it suffices to show
  \[ \left\{ g : \prod_{b \in B} \sum_{a \in f^{-1}[\{b\}]} P(a) \to Q(b) \right\} \cong \left\{ h :
  \prod_{a \in A} P(a) \to Q(f(a)) \right\} \]
  satisfying naturality. It is straightforward to check that $\iota$ (mapping from left to right)
  and its putative inverse defined below indeed form such an isomorphism between these sets:
  \begin{align*}
    \iota(g) &= a \mapsto p \mapsto g(f(a))(a, p)\\
    \iota^{-1}(h) &= b \mapsto (a, p) \mapsto h(a)(p)
    \qedhere
  \end{align*}
  % Note that the inverse definition type-checks because $h(a)(p) \in Q(f(a)) = Q(b)$ since $a \in
  % f^{-1}[\{b\}]$.
\end{proof}

We will also make use of a lemma that characterizes the morphisms into a reindexed family: they are
isomorphic to those morphisms into the original family that mapped indexing sets according to $f$.

\begin{lemma}
  \label{lem:famset-morphism-iso}
  Let $(A, P)$ and $(B, Q)$ be objects in $\FamSet$. Let $f : A \to B$ be a function. Then there is
  an isomorphism between the set of $\FamSet$-morphisms $h : (A, P) \to (B, Q)$ such that $Uh = f$
  and $\FamSet_A$ morphisms $(A, P) \to f^*(B, Q)$:
  \begin{align*}
  &\{ h \in \FamSet((A, P), (B, Q)) \mid Uh = f \} \cong\\
  &\FamSet_A((A, P), f^*(B, Q))
  \end{align*}
\end{lemma}

\begin{exercise}
  \label{ex:famset-morphism-iso}
  Prove Lemma \ref{lem:famset-morphism-iso}.
\end{exercise}

We can now update Figure \ref{fig:famset-diag-no-adjoints} to summarize these additional results:

\begin{center}
\begin{tikzpicture}
  \node (FamSet) at (0, 1.75) {};
  \node (Set) at (0, -1.5) {};
	\draw[draw=black, thin, solid] (FamSet) ellipse (3.75 and 1.75);
	\draw[draw=black, thin, solid] (Set) ellipse (1.50 and 0.50);
	\node[black, anchor=south west] at (1.3,-1.3) {$\Set$};
	\node[black, anchor=south west] at (1.5,3.25) {$\FamSet$};
	\node[black, anchor=south] at (-1.75,2.4) {$\FamSet_A$};
	\node[black, anchor=south] at (1.75,2.4) {$\FamSet_B$};
  \node (FamSetA) at (-1.75, 1.5) {};
  \node (FamSetB) at (1.75, 1.5) {};
	\draw[draw=black, thin, solid] (FamSetA) ellipse (1 and 1);
	\draw[draw=black, thin, solid] (FamSetB) ellipse (1 and 1);
	\draw[draw=black, thin, dashed] (-0.5,-1.3) -- (-1.5,0.50);
	\draw[draw=black, thin, dashed] (0.5,-1.3) -- (1.5,0.50);
	\node[black] (A) at (-0.5,-1.5) {$A$};
	\node[black] (B) at (0.5,-1.5) {$B$};
	\node[black] at (-1.75,1.75) {$(A, P)$};
	\node[black] at (-1.75,1.25) {$(A, P')$};
	\node[black] at (1.75,1.75) {$(B, Q)$};
	\node[black] at (1.75,1.25) {$(B, Q')$};
  \draw[draw=black, ->] (A) -- (B) node [midway, below=2pt, label=$f$] {};
  \draw[draw=black, ->] (FamSetA)++(1,0.5) -- (0.75, 2) node [midway, below=2pt, label=$\Sigma_f$] {}; 
  \draw[draw=black, <-] (FamSetA)++(1,-0.5) -- (0.75, 1) node [midway, below=2pt, label=$f^*$] {};
  \node[black] at (0, 1.7) {$\bot$};
  \draw[draw=black, <-] (Set)++(-1,0.5) -- (-1, 0) node [midway, below left=6pt, label=$U$] {};
  \draw[draw=black, ->] (Set)++(0,0.5) -- (0, 0) node [midway, below left=6pt, label=$\top$] {};
  \draw[draw=black, <-] (Set)++(1,0.5) -- (1, 0) node [midway, below right=10pt, label=$\{-\}$] {};
  \node[black] at (-0.7, -0.4) {$\dashv$};
  \node[black] at (0.5, -0.4) {$\dashv$};
\end{tikzpicture}
\captionof{figure}{A revision of Figure \ref{fig:famset-diag-no-adjoints}, now with adjoint relationships and
the reindexing functor ($\pi$ has been removed for clarity).}\label{fig:famset-diag-adjoints}
\end{center}

\subsection{A Proof of Correctness}

We want to show that, given an endofunctor $F$ on $\Set$ with an initial algebra, the endofunctor
$F^\alpha$ on $\FamSet_A$ has one as well: that is, we must find an initial object in
$\Alg_{F^\alpha}$. Since we already have an initial object in $\Alg_F$, it will be useful to relate
$\Alg_{F^\alpha}$ to this category. Mirroring our construction of $F^\alpha$ itself, we will do this
in two stages: first, we will consider algebras over $\hat F : \FamSet \to \FamSet$, which we relate
to algebras over $F$; then, we will limit our attention to an ``$\alpha$-based restriction'' of
$\hat F$-algebras.

Algebras over $\hat F$ are elements of the category $\Alg_{\hat F}$. Since $\FamSet$ ``sits atop''
$\Set$ (in the sense that a family $P$ is ``atop'' its indexing set $UP$), the category $\Alg_{\hat
F}$ likewise ``sits atop'' $\Alg_F$, a fact witnessed by the following \emph{algebras
fibration}.

\begin{definition}[Algebras fibration]
  The \emph{algebras fibration} $U^\Alg$ induced by the families fibration $U$ is a functor
  $U^{\Alg} : \Alg_{\hat F} \to \Alg_F$ defined by
  \begin{align*}
    U^\Alg(P, k : \hat F P \to P) &= (UP, Uk)\\
    U^\Alg(f : (P, k) \to (Q, k')) &= Uf : (UP, Uk) \to (UQ, Uk')
  \end{align*}
\end{definition}

It is worth pausing to type-check this definition, in which we have taken some notational shortcuts
for concision. Recall that $U : \FamSet \to \Set$ maps families to their indexing sets. Since $\hat
F$ is an endofunctor on \emph{families}, $P$ comprises a family indexed by some set; $k$ is a
morphism of families, which is a pair of functions; and $f$ is an algebra morphism (as defined in
Definition \ref{def:cat-of-algebras}) over an endofunctor on families and thus also a morphism of
families (satisfying a certain commutativity condition). Thus, $U$ has an action on each of these.

The algebras over $F$ inherit an additional useful property from those over $F$: in particular, if
$F$ has an initial algebra, than so does $\hat F$:

\begin{theorem}
  \label{thm:lifting-init-alg}
  Suppose $F$ has an initial algebra. Then $\hat F$ has an initial algebra too, and its carrier is
  $\top(\mu F)$.
\end{theorem}
\begin{proof}
  The key insight for this proof is that $F$-algebras with carrier $\{(A, P)\}$ are interderivable
  with $\hat F$-algebras with carrier $\{(A, P)\}$. A proof is provided by Ghani et
  al.~\citep{ghani2010fibrational}; we will not give one here. Using this fact, the result follows
  almost immediately: the carrier $\mu F$ of the initial $F$-algebra is isomorphic to $\sum_{t \in
  \mu F}\{(\_ \mapsto \star)\}$ as a set, since the summand is a singleton. This set is precisely
  $\{\top(\mu F)\}$, so that $\top(\mu F)$ is the carrier of the initial $\hat F$-algebra by the
  aforementioned interderivability fact.
\end{proof}

However, since $F^\alpha$ is a restriction of $\hat F$ by values of an algebra $(A, \alpha)$, we
want to look at a correspondingly restricted category of algebras, not $\Alg_{\hat F}$ in its
entirety. We thus restrict the category $\Alg_{\hat F}$ to an appropriate \emph{fiber} over our
refining algebra. This definition, given below, is directly analogous to the definition of fibers
over $\FamSet$; however, we phrase it slightly differently to suggest the generality of such a
construction. (One may find it worthwhile to ``pattern-match'' the tuple-valued variables below and
unfold the definitions of $U^\Alg$ and $U$ to see this correspondence more directly.) Note that we
are also now adopting the convention of abbreviating $(A, \alpha)$ as simply $\alpha$.

\begin{definition}[Fiber of the algebras fibration over $\alpha$]
  The \emph{fiber} of the algebras fibration over $\alpha$, or equivalently the \emph{category of
  $\hat F$-algebras over $\alpha$}, denoted by $(\Alg_{\hat F})_\alpha$, is defined as follows:
  \begin{itemize}
    \item Objects are $\hat F$-algebras $(P, k)$ such that $U^\Alg (P, k) = \alpha$.
    \item Morphisms are $\hat F$-algebra morphisms $f : (P, k) \to (Q, k')$ such that $U f =
    \mathsf{id}_{\alpha}$.
  \end{itemize}
\end{definition}

Once again, it is worth briefly type-checking this definition. For instance, the criterion on
objects is well defined: $k = (k_1, k_2)$ is a morphism $\hat F P \to P$, which is an abbreviation
of $\hat F (A, P) \to (A, P)$ with the indexing set explicit. By the definition of $\FamSet$, $k_1$
maps between the carriers; since $\hat F (A, P)$ has carrier $F A$ by definition, we thus have $k_1
: F A \to A$. And since $U^\Alg (P, k) = U^\Alg ((A, P), (k_1, k_2)) = (A, k_1)$, this is of the
same type as $\alpha$, i.e., as $(A, \alpha : F A \to A)$.

Moreover, it turns out that $(\Alg_{\hat F})_\alpha$ is the same category as $\Alg_{F^\alpha}$, just
described as a ``decomposition'' in terms of $\hat F$ and $\alpha$.

\begin{lemma}
  \label{lem:alg-falpha-iso}
  Let $(A, \alpha)$ be an $F$-algebra. Then there is an isomorphism of categories
  \[ (\Alg_{\hat F})_\alpha \cong \Alg_{F^\alpha} \]
\end{lemma}

\begin{proof}
  Observe that an object of $(\Alg_{\hat F})_\alpha$ is a pair of a family and a
  $\FamSet$-morphism---i.e., a tuple $((A, P), k : \hat F(A, P) \to (A, P))$---such that $Uk =
  \alpha$. By Lemma \ref{lem:famset-morphism-iso}, we can characterize such a $\FamSet$-morphism as
  a $\FamSet_{FA}$-morphism via the reindexing functor, so the object is equivalently expressed as
  $((A, P), k' : \hat F(A, P) \to \alpha^*(A, P))$. In turn, we observe that morphisms $k' : \hat
  F(A, P) \to \alpha^*(A, P)$ are themselves in isomorphism with $\FamSet_A$-morphisms $k'' :
  \Sigma_\alpha \hat F(A, P) \to (A, P)$ by Lemma \ref{lem:reindex-adjoint}. Recalling the
  definition of $F^\alpha$, we can rewrite this as $k'' : F^\alpha(A, P) \to (A, P)$. Thus, we can
  again equivalently rewrite our object as $((A, P), k'' : F^\alpha(A, P) \to (A, P))$, which is
  precisely an object in $\Alg_{F^\alpha}$. Transporting a morphism along the corresponding sequence
  of isomorphic transformations yields the required action on morphisms.
\end{proof}

In order to work with this new formulation of the category, we will reproduce for $(\Alg_{\hat
F})_\alpha$ several results analogous to those we showed for the families fibration. The following
lemmas are analogues of Lemmas \ref{lem:reindex-adjoint} and \ref{lem:famset-morphism-iso}; their
proofs are technical but not exceedingly difficult, so we elide them.

\begin{lemma}
\label{lem:opreindex-alg-adjoint}
  Let $f$ be an $F$-algebra morphism $(A, \alpha) \to (B, \beta)$. Then there are functors
  $\Sigma_f^\Alg : (\Alg_{\hat F})_\alpha \to (\Alg_{\hat F})_\beta$ and $f^{*\Alg} : (\Alg_{\hat
  F})_\beta \to (\Alg_{\hat F})_\alpha$ such that $\Sigma_f^\Alg \dashv f^{*\Alg}$ and for any $\hat
  F$-algebras $((A, P), k)$ and $((B, Q), k')$,
  \begin{itemize}
    \item The $\hat F$-algebra $\Sigma_f^{\Alg}((A, P), k)$ has carrier $\Sigma_f(A, P)$, and
    \item The $\hat F$-algebra $f^{*\Alg}((B, Q), k')$ has carrier $f^*(B, Q)$.
  \end{itemize}
\end{lemma}

\begin{lemma}
\label{lem:alg-hom-iso-reindex}
  Let $(A, \alpha)$ and $(B, \beta)$ be $\hat F$-algebras. Fix an object $((A, P), k)$ in
  $(\Alg_{\hat F})_\alpha$ and $((B, Q), k')$ in $(\Alg_{\hat F})_\beta$. Let $f : (A, \alpha) \to
  (B, \beta)$ be an $\hat F$-algebra morphism. Then there is an isomorphism between the set of $\Alg_{\hat F}$-morphisms $h : k \to k'$ such that
  $U^\Alg h = f$ and the set of morphisms $k \to f^{*\Alg}(k')$:
  \[ \{h \in \Alg_{\hat F}(k, k') \mid U^\Alg h = f\} \cong \Alg_{\hat F}(k, f^{*\Alg}(k')). \]
\end{lemma}

We can now draw a diagram analogous to Figure \ref{fig:famset-diag-adjoints} for the algebras
fibration. Note that we temporarily drop our abbreviated notation for families and algebras to
better analogize this diagram to the one for the families fibration.

\begin{center}
\begin{tikzpicture}
  \node (FamSet) at (0, 1.75) {};
  \node (Set) at (0, -1.5) {};
	\draw[draw=black, thin, solid] (FamSet) ellipse (3.75 and 1.75);
	\draw[draw=black, thin, solid] (Set) ellipse (1.50 and 0.50);
	\node[black, anchor=south west] at (1.3,-1.3) {$\Alg_{F}$};
	\node[black, anchor=south west] at (1.5,3.25) {$\Alg_{\hat F}$};
	\node[black, anchor=south] at (-1.75,2.4) {$(\Alg_{\hat F})_\alpha$};
	\node[black, anchor=south] at (1.75,2.4) {$(\Alg_{\hat F})_\beta$};
  \node (FamSetA) at (-1.75, 1.5) {};
  \node (FamSetB) at (1.75, 1.5) {};
	\draw[draw=black, thin, solid] (FamSetA) ellipse (1.25 and 1);
	\draw[draw=black, thin, solid] (FamSetB) ellipse (1.25 and 1);
	\draw[draw=black, thin, dashed] (-0.75,-1.3) -- (-1.5,0.50);
	\draw[draw=black, thin, dashed] (0.75,-1.3) -- (1.5,0.50);
	\node[black] (A) at (-0.75,-1.5) {$(A, \alpha)$};
	\node[black] (B) at (0.75,-1.5) {$(B, \beta)$};
	\node[black] at (-1.75,1.75) {$((A, P), (\alpha, k))$};
	\node[black] at (-1.75,1.25) {$((A, P'), (\alpha, k'))$};
	\node[black] at (1.75,1.75) {$((B, Q), (\beta, j))$};
	\node[black] at (1.75,1.25) {$((B, Q'), (\beta, j'))$};
  \draw[draw=black, ->] (A) -- (B) node [midway, below=2pt, label=$f$] {};
  \draw[draw=black, ->] (FamSetA)++(1.2,0.5) -- (0.55, 2) node [midway, below=2pt, label=$\Sigma_f^\Alg$] {}; 
  \draw[draw=black, <-] (FamSetA)++(1.2,-0.5) -- (0.55, 1) node [midway, below=2pt, label=$f^{*\Alg}$] {};
  \node[black] at (0, 1.7) {$\bot$};
  \draw[draw=black, <-] (Set)++(0,0.5) -- (0, 0) node [midway, below left=10pt, label=$U^\Alg$] {};
\end{tikzpicture}
\captionof{figure}{A summary of constructions for the algebras fibration.}\label{fig:alg-diag-adjoints}
\end{center}

Equipped with these intermediate results, we can now finally prove what we set out to show: that
$F^\alpha$ admits an initial algebra, and that its carrier is isomorphic to the ``na\"ive''
refinement interpretation. We make one brief observation: we can rewrite our previous ``na\"ive''
interpretation of a refinement by $\llpr{\alpha} : \mu F \to A$ using the truth and op-reindexing
functors as follows:

\[ (A, a \mapsto \{x \in \mu F \mid \llpr{\alpha}(x) = a\}) \cong \Sigma_{\llpr{\alpha}} \top(\mu F). \]

Using this formulation, we now state our main theorem.

\begin{theorem}
The endofunctor $F^\alpha$ has an initial algebra, and its carrier is $\Sigma_{\llpr{\alpha}}
\top(\mu F)$.
\end{theorem}

\begin{proof}
  We will first construct a candidate algebra in $(\Alg_{\hat F})_\alpha$, then show that it is
  initial. Note that, by Lemma \ref{lem:alg-falpha-iso}, this is equivalent to identifying an
  $F^\alpha$-algebra and showing its initiality in $\Alg_{F^\alpha}$.
  
  From Theorem \ref{thm:lifting-init-alg}, we obtain an inital algebra $(\top(\mu F),
  \mathsf{in}_{\hat F})$ in $\Alg_{\hat F}$. We will use this to construct an initial object in the
  fiber over $\alpha$. Mirroring the construction of $F^\alpha$, we apply the op-reindexing functor
  $\Sigma_{\llpr{\alpha}}^\Alg$ for the algebras fibration to obtain
  \[ \left(\Sigma_{\llpr{\alpha}}^\Alg \top(\mu F),~ \Sigma_{\llpr{\alpha}}^\Alg \mathsf{in}_{\hat
     F} : \hat F (\top(\mu F)) \to \top(\mu F) \right). \]
  We will aim to show that this object is our desired initial object in $(\Alg_{\hat F})_\alpha$.

  First, observe that by Lemma \ref{lem:opreindex-alg-adjoint}, we have
  \[ \Sigma_{\llpr{\alpha}}^\Alg \top(\mu F) \cong \Sigma_\llpr{\alpha} \top(\mu F), \]
  i.e., the carrier of this algebra is our na\"ive refinement encoding. So it remains only to show
  that the algebra is indeed initial.

  We show initiality directly from the definition: we show that there is exactly one morphism
  between $\Sigma_{\llpr{\alpha}}^\Alg \mathsf{in}_{\hat F}$ and an arbitrary $\hat F$-algebra $k :
  \hat F(A, P) \to (A, P)$ (hereafter we will denote each algebra under discussion by its structure
  map alone). We proceed by considering a series of isomorphisms of hom-sets:
  \begin{align*}
    &(\Alg_{\hat F})_\alpha\left(\Sigma_{\llpr{\alpha}}^\Alg \mathsf{in}_{\hat F},~ k\right)\\
    &\cong (\Alg_{\hat F})_{\mathsf{in}_F}\left(\mathsf{in}_{\hat F},~ \llpr{\alpha}^*(k)\right) \tag{Adjointness portion of Lemma \ref{lem:opreindex-alg-adjoint}}\\
    &\cong \{ h \in \Alg_{\hat F}(in_{\hat F}, k) \mid U^\Alg h = \llpr{\alpha} \} \tag{Lemma \ref{lem:alg-hom-iso-reindex}}
  \end{align*}
  Notice that this final set is a subset of morphisms out of an initial object, so it can have at
  most one element (namely, $\llpr{k}$). If this is to be a singleton set, then it must indeed
  contain $\llpr{k}$, so in particular it must be that
  \[ U^\Alg \llpr{k} = \llpr{\alpha}. \]
  Recall that $U^\Alg$ is left adjoint, so it preserves colimits: the above equation (of morphisms
  from initial objects) thus holds.

  Accordingly, the algebra $\Sigma_{\llpr{\alpha}}^\Alg \mathsf{in}_{\hat F}$ is indeed initial in
  $(\Alg_{\hat F})_\alpha$, completing the proof.
\end{proof}

Before concluding this section, it is worth noting that the construction given here can be extended
in various ways, such as to support refinements defined in terms of functions that are not directly
expressed as folds. We direct the reader to Atkey et al.~\citep{atkey2011refining} for details.

\section{A (Literal) Invitation to Fibered Category Theory}

Throughout this chapter, we have followed Atkey et al.~\citep{atkey2011refining} in carefully
presenting the material so as to admit a generalization to the abstract language of
\emph{fibrations}, and have gradually adjusted our presentation over the course of the chapter
toward a more abstract framing in terms of fibrational constructions. Regrettably, we do not have
the space here to present the full generalization---a construction that can be carried out in any
\emph{full Cartesian Lawvere category with very strong coproducts}---or to discuss fibrations in
their own right.

Instead, we direct the reader to the following sources. The original development of this material by
Atkey et al.~\citep{atkey2011refining} gestures further toward its generalization, in addition to
providing several further constructions that translate broader classes of refinements into indexed
inductive types. Ghani et al.'s paper~\citep{ghani2010fibrational} includes the original, abstract
derivation of the truth-preserving lifting $\hat F$ we employ in this paper. Some relevant
background can also be found in Jacobs' paper on comprehension
categories~\citep{jacobs1993comprehension}, while the same author's textbook \emph{Categorical Logic
and Type Theory}~\citep{jacobs1999categorical} provides an exposition of fibered category theory and
several applications. A relatively limited selection of reading from among these sources should
provide sufficient background to interpret this chapter in a general fibrational setting, thereby
characterizing the precise categorical structure required to unify refinement and indexed types.

\newpage
\section*{Solutions to Exercises}

\soltitle{ex:denote-bool}

Pick $\llbr{\texttt{true}} = \inl$ and $\llbr{\texttt{false}} = \inr$, respectively the left and right
injections into the coproduct, and take $\llbr{\texttt{if}(t, e)} = \langle t, e \rangle$.
Commutativity and recursor uniqueness are immediate from the universal property for coproducts.

\soltitle{ex:interp-exnat}

Since the recursor for \texttt{ExtNat} is not explicitly given, we first derive it:

\begin{mathpar}
\inferrule*
  {\vdash e_\texttt{z} : \tau \\ x : \tau \vdash e_\texttt{s} : \tau \\ \vdash e_\texttt{i} : \tau}
  {\vdash \texttt{rec}(e_\texttt{z}, x.e_\texttt{s}, e_\texttt{i}) : \tau}
\end{mathpar}

As the hint alludes to, this example differs from the preceding ones because there are three
constructors rather than two, yielding a third square in our diagram; it is a ``revolving door''
rather than a pair of squares. It looks as follows:

% https://q.uiver.app/#q=WzAsOCxbMiwyLCJcXGxsYnJ7RXh0TmF0fSJdLFs0LDIsIlxcbGxicntFeHROYXR9Il0sWzAsNCwiMSJdLFswLDYsIjEiXSxbMiw0LCJcXGxsYnJ7XFx0YXV9Il0sWzQsNCwiXFxsbGJye1xcdGF1fSJdLFswLDAsIjEiXSxbMCwyLCIxIl0sWzEsMCwiXFx0ZXh0dHR7c3VjY30iLDJdLFsyLDAsIlxcdGV4dHR0e3plcm99IiwwLHsibGFiZWxfcG9zaXRpb24iOjIwfV0sWzIsMywiXFxpZF8xIiwyXSxbMyw0LCJlX1xcdGV4dHR0e3p9IiwyXSxbMCw0LCJcXHRleHR0dHtyZWN9X1xcdGV4dHR0e0V4dE5hdH0oZV9cXHRleHR0dHt6fSwgZV9cXHRleHR0dHtpfSwgeC5lX1xcdGV4dHR0e3N9KSJdLFsxLDUsIlxcdGV4dHR0e3JlY31fXFx0ZXh0dHR7RXh0TmF0fShlX1xcdGV4dHR0e3p9LCBlX1xcdGV4dHR0e2l9LCB4LmVfXFx0ZXh0dHR7c30pIl0sWzUsNCwiZV9cXHRleHR0dHtzfSJdLFs2LDAsIlxcdGV4dHR0e2luZn0iXSxbNiw3LCJcXGlkXzEiLDJdLFs3LDQsImVfXFx0ZXh0dHR7aX0iLDAseyJsYWJlbF9wb3NpdGlvbiI6MzB9XV0=
\[\begin{tikzcd}
	1 \\
	\\
	1 && {\llbr{\texttt{ExtNat}}} && {\llbr{\texttt{ExtNat}}} \\
	\\
	1 && {\llbr{\tau}} && {\llbr{\tau}} \\
	\\
	1
	\arrow["{\id_1}"', from=1-1, to=3-1]
	\arrow["{\texttt{inf}}", from=1-1, to=3-3]
	\arrow["{e_\texttt{i}}"{pos=0.25}, from=3-1, to=5-3]
	\arrow["{\texttt{rec}_\texttt{ExtNat}(e_\texttt{z}, x.e_\texttt{s}, e_\texttt{i})}", from=3-3, to=5-3, dashed]
	\arrow["{\texttt{succ}}"', from=3-5, to=3-3]
	\arrow["{\texttt{rec}_\texttt{ExtNat}(e_\texttt{z}, x.e_\texttt{s}, e_\texttt{i})}", from=3-5, to=5-5, dashed]
	\arrow["{\texttt{zero}}"{pos=0.3}, from=5-1, to=3-3]
	\arrow["{\id_1}"', from=5-1, to=7-1]
	\arrow["{e_\texttt{s}}", from=5-5, to=5-3]
	\arrow["{e_\texttt{z}}"', from=7-1, to=5-3]
\end{tikzcd}\]

\soltitle{ex:self-trees}

\begin{enumerate}
  \item The actions of $F_{\texttt{SelfTree}}$ on objects and morphisms are given below:
  \begin{align*}
    F_{\texttt{SelfTree}}(X) &= 1 \oplus A \times X^A\\
    F_{\texttt{SelfTree}}(e) &= [\mathsf{id}_1, \langle \mathsf{id}_A \circ \pi_2, (e \circ -) \circ \pi_1\rangle]
  \end{align*}
  \item The relevant algebra is
  \[ \alpha_{\texttt{selfHeight}} : F_{\texttt{SelfTree}} \llbr{\texttt{Nat}} \to \llbr{\texttt{Nat}} = [0, \texttt{succ} \circ \mathsf{app} \circ \mathsf{flip} ] \]
  where $\mathsf{flip}$ is the isomorphism that witnesses the commutativity of products.
\end{enumerate}

\soltitle{ex:app-as-algebra}

Fix objects $A, B$ and consider their exponential. Since $\texttt{app} : B^A \times A \to B$, we are
led to define the endofunctor $F(X) = X^A \times A$, whose action on morphisms is given by $F(e) =
\langle (e \circ -) \circ \pi_1, \mathsf{id}_A \circ \pi_2\rangle$. (Notice that, despite the type
of \texttt{app} containing two objects, there is only one possible choice of endofunctor: if we
tried to define the functor in $A$, it would be contravariant.)

We claim that the identity morphism for the initial object $0$ is the initial $F$-algebra. To see
why, first observe that $F(0) = 0^A \times A \cong 0 \times A \cong 0$, so the choice of initial
algebra $\mathsf{id}_0 : F(0) \to 0$ (transported across the relevant isomorphisms) type-checks.
Moreover, its initiality is immediate from the fact that $0$ is initial.

By Lambek's lemma, we thus have that $\mu F \cong 0$, so the inductive type that this functor
characterizes is empty. We can read off the corresponding \texttt{inductive} definition:

\begin{lstlisting}
inductive Empty' (A : Type) : Type where
  | mk : (A -> Empty' A) -> Empty' A -> Empty' A
\end{lstlisting}

\soltitle{ex:isorec-translation}

We translate each rule in turn.

\begin{itemize}
  \item \textsc{T-fold}: As previously discussed, the positivity hypothesis of this rule is
        analogous to the requirement that the endofunctor $F$ denoted by $t. \tau$ has an initial
        algebra. The substitution $\tau[\mu t.\tau/t]$ corresponds to plugging in $\llbr{\tau}$ for
        the functor variable $t$ in $F(t)$: it is functor action on objects. Thus, the conclusion of
        the rule induces the following morphism:

        % https://q.uiver.app/#q=WzAsMixbMCwwLCJcXGxsYnJ7RiAoXFxtdSBGKX0iXSxbMiwwLCJcXGxsYnJ7XFxtdSBGfSJdLFswLDEsIlxcdGV4dHR0e2ZvbGR9Il1d
        \[\begin{tikzcd}
          {\llbr{F (\mu F)}} && {\llbr{\mu F}}
          \arrow["{\texttt{fold}}", from=1-1, to=1-3]
        \end{tikzcd}\]
  \item \textsc{T-rec}: Now that we know how to interpret type-variable substitution, this
        translation proceeds mechanically: the hypothesis says we have a morphism $e_1 : \llbr{\rho}
        \to F\llbr{\rho}$, and the conclusion uses this to produce a morphism $\mu F \to
        \llbr{\rho}$:

        % https://q.uiver.app/#q=WzAsMyxbMiwwLCJcXGxsYnJ7XFxtdSBGfSJdLFswLDIsIkYgXFxsbGJye1xccmhvfSJdLFsyLDIsIlxcbGxicntcXHJob30iXSxbMSwyLCJlXzEiXSxbMCwyLCJcXHRleHR0dHtyZWN9KHguZV8xKSIsMV1d
        \[\begin{tikzcd}
          && {\llbr{\mu F}} \\
          \\
          {F \llbr{\rho}} && {\llbr{\rho}}
          \arrow["{\texttt{rec}(x.e_1)}"{description}, from=1-3, to=3-3]
          \arrow["{e_1}", from=3-1, to=3-3]
        \end{tikzcd}\]
  \item \textsc{$\beta^\texttt{rec}$}: Now we must contend with the function $\texttt{map}(e)$.
        Observe that $e : (t.\tau)[\rho/t]$ from rule \textsc{T-map}, so we know that the type of
        $e$ denotes a functor application $F \llbr{\rho}$. The key insight---which is most easily
        observed by inspecting the sample $\beta$ rules for this function---is that the function
        $\texttt{map}(e)$ denotes precisely the action of $F$ on \emph{morphisms}, i.e., terms.
        Thus, we translate $\texttt{map}(e)$ as the morphism action $F(-)$.

        We can now interpret the $\beta^\texttt{rec}$ rule. The left-hand side corresponds to the
        morphism $\texttt{rec}(x.e_1) \circ \texttt{fold}$. The right-hand side becomes (removing
        the explicit point $y$ for clarity, and recalling that substitution becomes morphism
        composition) $e_1 \circ \llbr{\texttt{map}(e_2)(\texttt{rec}(x. e_1))}$, which
        unfolds to $e_1 \circ F(\texttt{rec}(x.e_1))$. Since this rule is equational, we have
        that these two morphisms are equal, i.e.,
        \[ \texttt{rec}(x.e_1) \circ \texttt{fold} = e_1 \circ F(\texttt{rec}(x.e_1)) \]
        This just says that if we stitch together the diagrams from the preceding two rules, the
        resulting square commutes:
        % https://q.uiver.app/#q=WzAsNCxbMiwwLCJcXGxsYnJ7XFxtdSBGfSJdLFswLDIsIkYgXFxsbGJye1xccmhvfSJdLFsyLDIsIlxcbGxicntcXHJob30iXSxbMCwwLCJGIChcXG11IEYpIl0sWzEsMiwiZV8xIl0sWzAsMiwiXFx0ZXh0dHR7cmVjfSh4LmVfMSkiLDFdLFszLDAsIlxcdGV4dHR0e2ZvbGR9IiwxXSxbMywxLCJcXHRleHR0dHtGfShcXHRleHR0dHtyZWN9KHguIGVfMSkpIiwxXV0=
        \[\begin{tikzcd}
          {F (\mu F)} && {\llbr{\mu F}} \\
          \\
          {F \llbr{\rho}} && {\llbr{\rho}}
          \arrow["{\texttt{fold}}"{description}, from=1-1, to=1-3]
          \arrow["{\texttt{F}(\texttt{rec}(x. e_1))}"{description}, from=1-1, to=3-1]
          \arrow["{\texttt{rec}(x.e_1)}"{description}, from=1-3, to=3-3]
          \arrow["{e_1}", from=3-1, to=3-3]
        \end{tikzcd}\]
  \item $\eta^\texttt{rec}$: The hypothesis of this rule translates to
        \[ e \circ \texttt{fold} = e_1 \circ F(\texttt{rec}(x.e_1)) \]
        and its conclusion to
        \[ e = \texttt{rec}(x.e_1). \]
        In other words, any morphism $e$ satisfying the equality from the $\beta$ rule must be equal
        to the recursor; that is, the recursor is unique. We reflect this in our diagram from above:
        \[\begin{tikzcd}
          {F (\mu F)} && {\llbr{\mu F}} \\
          \\
          {F \llbr{\rho}} && {\llbr{\rho}}
          \arrow["{\texttt{fold}}"{description}, from=1-1, to=1-3]
          \arrow["{\texttt{F}(\texttt{rec}(x. e_1))}"{description}, from=1-1, to=3-1]
          \arrow["{\texttt{rec}(x.e_1)}"{description}, from=1-3, to=3-3, dashed]
          \arrow["{e_1}", from=3-1, to=3-3]
        \end{tikzcd}\]
\end{itemize}
Observe that (renaming \texttt{fold} to $c$, $e_1$ to $f_c$, and $\rho$ to $A$) we have now arrived
at the same diagram we derived in our previous translation!

\soltitle{ex:fin-functor}

\[ F_{\texttt{Fin}}(\N, P) = \left(\N, \begin{cases}0 \mapsto \varnothing \\ \texttt{succ}(n) \mapsto
  1 \oplus P(n)
\end{cases}\right) \]

\soltitle{ex:very-strong-coprods}

First, note that the ``straightforward'' morphism $\psi$ is concretely
\[ \psi(a, p) = (f, a \mapsto p \mapsto (a, p)). \]

(Note that we are using the isomorphic formulation of indexed coproducts given in Equation
(\ref{eqn:indexed-coprod-simplified}) to minimize bureaucracy.)

\begin{enumerate}
  \item Unfolding definitions, we find that
  \[ (f \circ \pi_{(A, P)})(a, p) = f(a) \]
  and
  \[ (\pi_{\Sigma_f(A, P)}\circ \{\psi\})(a, p) = \pi_{\Sigma_f(A, P)}(f(a), a, p) = f(a) \]
  so that
  \[ f \circ \pi_{(A, P)} = \pi_{\Sigma_f(A, P)}\circ \{\psi\} \]
  as required.

  \item Define the putative inverse of $\{\psi\}$ to be $\iota : \{\Sigma_f(A, P)\} \to \{(A, P)\}$
  given by
  \[ \iota(b, (a, p)) = (a, p) \]

  Then observe that
  \[ \{\psi\}(\iota(b, (a, p))) = \{\psi\}(a, p) = (f(a), (a, p)) = (b, (a, p)) \]
  where the last equality follows by the definition of $\Sigma_f(A, P)$, and moreover
  \[ \iota(\{\psi\}(a, p)) = \iota(f(a), (a, p)) = (a, p) \]
  so that $\{\psi\}$ is indeed an isomorphism with inverse $\iota$.
\end{enumerate}

\soltitle{ex:trivial-truth-pres}

This is clearly a lifting, since
\[ F(U(A, P)) = F A = U(FA, \_ \mapsto 1) = U(\hat F (A, P)) \]
and
\[ F(U(f, f^\sim)) = F f = U(f, \_ \mapsto \mathsf{id}_1) = U(\hat F(f, f^\sim)). \]

Moreover, that it is truth preserving is witnessed by the natural isomorphism
\[ \eta_{(A, P)} = \mathsf{id}_{(A, P)}  \]
since, noting that
\[ \top(F(A)) = (FA, \_ \mapsto 1) = \hat F(\top(A)) \]
and
\[ \top(F(f)) = (f, \_ \mapsto \mathsf{id}_1) = \hat F(\top(f)) \]
the following square trivially commutes:
% https://q.uiver.app/#q=WzAsNCxbMCwwLCIoXFx0b3AgXFxjaXJjIEYpKEEpIl0sWzIsMCwiKFxcaGF0IEYgXFxjaXJjIFxcdG9wKShBKSJdLFswLDIsIihcXHRvcCBcXGNpcmMgRikoQikiXSxbMiwyLCIoXFxoYXQgRiBcXGNpcmMgXFx0b3ApKEIpIl0sWzAsMSwiXFxldGFfeyhBLCBcXF9cXG1hcHN0byAxKX0iXSxbMiwzLCJcXGV0YV97KEIsIFxcXyBcXG1hcHN0byAxKX0iXSxbMCwyLCIoXFx0b3AgXFxjaXJjIEYpKGYpIiwxXSxbMSwzLCIoXFxoYXQgRiBcXGNpcmMgXFx0b3ApKGYpIiwxXV0=
\[\begin{tikzcd}
	{(\top \circ F)(A)} && {(\hat F \circ \top)(A)} \\
	\\
	{(\top \circ F)(B)} && {(\hat F \circ \top)(B)}
	\arrow["{\eta_{(A, \_\mapsto 1)}}", from=1-1, to=1-3]
	\arrow["{(\top \circ F)(f)}"{description}, from=1-1, to=3-1]
	\arrow["{(\hat F \circ \top)(f)}"{description}, from=1-3, to=3-3]
	\arrow["{\eta_{(B, \_ \mapsto 1)}}", from=3-1, to=3-3]
\end{tikzcd}\]

\soltitle{ex:bh-indexed-rbtrees}

\begin{enumerate}
  \item
    \begin{align*}
      F_{\texttt{RBTree}}(X) &= 1 \oplus ((A \times (X \times X)) \oplus (A \times (X \times X)))\\
      F_{\texttt{RBTree}}(e) &= [\inl \circ \mathsf{id}_1, \inl \circ \inr \circ m, \inr \circ \inr \circ m ]\\
      &\text{where } m = \langle\mathsf{id}_A \circ \pi_1 \circ \pi_1, \langle e \circ \pi_1 \circ \pi_2, e \circ \pi_2 \circ \pi_2\rangle\rangle
    \end{align*}

  \item
  \[ \alpha_{\texttt{bh}} = [0, \texttt{succ} \circ \texttt{max} \circ \pi_2, \texttt{max} \circ \pi_2] \]

  \item 
  \begin{align*}
    &F^\texttt{bh}_\texttt{RBTree}(\N, P)\\
    &\cong (\N, n \mapsto \{y \in F_\texttt{RBTree}(\N, P) \mid \texttt{bh}((F_\texttt{RBTree}\pi_{(\N, P)})(y)) = n\})\\
    &= (\N, n \mapsto \{y \in 1 \oplus ((A \times ((\N, P) \times (\N, P))) \oplus (A \times ((\N, P) \times (\N, P)))) \mid\\
    &\hspace{6em} (y = \inl~() \text{ and } 0 = n)\text{ or}\\
    &\hspace{6em} (y = \inr~(\inl~(a, ((m_\ell, p_\ell), (m_r, p_r)))) \text{ and } n = \texttt{succ}(\texttt{max}(n_\ell, n_r)))\\
    &\hspace{6em} (y = \inr~(\inr~(a, ((m_\ell, p_\ell), (m_r, p_r)))) \text{ and } n = \texttt{max}(n_\ell, n_r))\})\\
    &= (\N, (n \mapsto
    \left(\begin{array}{ll}1 & \text{if } n = 0 \\ \varnothing & \text{else} \end{array}\right)
    \oplus\\
    &\hspace{5.75em}(\left(\begin{array}{ll} A \times ((\{m_\ell\} \times P(m_\ell)) \times (\{m_r\} \times P(m_r))) & \text{if } n = \texttt{succ}(\texttt{max}(m_\ell, m_r)) \\ \varnothing & \text{else}\end{array} \right)\oplus\\
    &\hspace{6.25em} \left(\begin{array}{ll} A \times ((\{m_\ell\} \times P(m_\ell)) \times (\{m_r\} \times P(m_r))) & \text{if } n = \texttt{max}(m_\ell, m_r) \\ \varnothing & \text{else} \end{array} \right))
  ))
  \end{align*}

  One can verify that this is isomorphic to the translation of the following inductive type:
\begin{lstlisting}
inductive BHRBTree : Nat -> Type where
  | leaf : BHRBTree 0
  | red  : {m n : Nat} -> A -> BHRBTree m -> BHRBTree n -> BHRBTree (succ (max m n))
  | red  : {m n : Nat} -> A -> BHRBTree m -> BHRBTree n -> BHRBTree (max m n)
\end{lstlisting}
\end{enumerate}

\soltitle{ex:famset-morphism-iso}

Recall that $\FamSet$-morphisms from $(A, P)$ to $(B, Q)$ are pairs 
\[\left(h : A \to B, h^\sim : \prod_{a \in A}P(a) \to Q(h(a))\right). \]
The left-hand set in the lemma statement is the restriction of such pairs to those whose first
element is $f$, so they are of the form
\[ \left(f, h^\sim : \prod_{a \in A}P(a) \to Q(f(a))\right) .\]

By the definitions of $\FamSet_A$ and $f^*$, morphisms in the right-hand set are pairs
\[ \left(\mathsf{id}_A, f^\sim : \prod_{a \in A} P(a) \to Q(f(a))\right). \]

Since the first component of every pair in either set is fixed, both sets are isomorphic to the
set of functions of type
\[ \prod_{a \in A} P(a) \to Q(f(a)) \]
and thus to each other.
